---
title: "Non-Gaussian random effect distributions"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Non-Gaussian random effect distributions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>",
                      message = FALSE, warning = FALSE)
## Every fit below is run for real and cached under inst/cache with
## nlmixr2save's `:=` operator, so this page shows live output without
## refitting on every build.  See vignettes/cacheSetup.R.
source("cacheSetup.R")
estCache("nonGaussianEta-")
```

A random effect in nlmixr2 is Normal by construction. Sometimes it should not
be: a heavy-tailed random effect is the standard robustness move against
influential individuals, and a clearance that is *itself* gamma distributed is
a different model from a log-normal one, not a reparameterization of it.

You declare the distribution in the `ini({})` block with a `dist()` line:

```{r model}
library(nlmixr2est)

gammaCl <- function() {
  ini({
    tka   <- 0.45
    lclm  <- log(1.1)   # gamma mean of CL
    lclrv <- log(0.09)  # gamma relative variance of CL
    tv    <- 3.45
    eta.ka ~ 0.6
    eta.cl ~ 1
    dist(eta.cl) ~ dgamma(shape = 1/exp(lclrv),
                          rate  = 1/(exp(lclrv)*exp(lclm)))
    add.sd <- 0.7
  })
  model({
    ka <- exp(tka + eta.ka)
    cl <- eta.cl        # a gamma distributed clearance
    v  <- exp(tv)
    d/dt(depot)  <- -ka*depot
    d/dt(center) <-  ka*depot - cl/v*center
    cp <- center/v
    cp ~ add(add.sd)
  })
}
```

Two things to notice. `eta.cl` **is** the clearance -- the declared
distribution is the marginal distribution of the random effect itself, so the
model block uses it directly rather than as a deviation from a typical value.
And its variance is written as `1`, not as a variance to estimate; the reason
is below.

`lotri::lotriEtaDists()` lists the families you can declare:

```{r families}
head(lotri::lotriEtaDists()[, c("name", "parNames", "support")], 12)
```

## How it works

The implementation is Bauer's (NONMEM 7.5.1): **keep the latent random effect
standard normal and change the CDF**.

```
z   ~ N(0, 1)          latent, unit variance, fixed
u   = phiU(z)          normal CDF        ->  U(0, 1)
eta = Q(u; args)       inverse CDF of the declared family
```

`rxEtaDistExpand()` performs that rewrite, and it is worth looking at once,
because everything downstream -- simulation and every estimation method -- sees
only its output:

```{r expand}
rxode2::rxEtaDistExpand(gammaCl())$lstExpr[1:3]
```

The declared random effect became an ordinary model variable computed from a
latent standard normal, and the arguments of `dgamma()` are ordinary `ini({})`
parameters, estimated like any others. That is the whole trick: **the
estimation machinery never needs a non-Gaussian prior**, because the random
effect it integrates over is still standard normal.

### Why the variance has to be one

Correlation between two declared random effects is induced on the *latent*
scale, through a Cholesky factor -- a Gaussian copula. That makes the block a
**correlation** matrix rather than a covariance one, so its diagonal has to be
one:

```r
eta.cl + eta.v ~ c(1,
                   0.5, 1)
dist(eta.cl) ~ dgamma(...)
dist(eta.v)  ~ dgamma(...)
```

`lotri()` says so if it is not. nlmixr2 cannot fix individual components of an
omega block, so this is stated by *estimating* the diagonal at one rather than
by fixing it, and the expansion turns the off-diagonals into unconstrained
`rxCor.*` parameters with a fixed identity omega -- a form every method already
handles. The fit reports the correlation back:
`fit$etaDistCor`, and `tanh()` of an `rxCor.*` estimate (its back-transformed
column) is that correlation directly.

## Fitting

Nothing special at the call site:

```{r fit}
fit := nlmixr2(gammaCl, nlmixr2data::theo_sd, est = "focei",
               control = foceiControl(print = 0))
```

```{r print}
print(fit)
```

The gamma's own arguments come back as ordinary population parameters, and the
fit carries both the latent normal and the gamma-distributed clearance per
subject:

```{r etas}
head(fit$eta)
head(fit[, c("ID", "rxz.eta.cl", "eta.cl")], 3)
```

## Which methods support this

The FOCEi family, `saem`, `posthoc` and simulation. A method that cannot use a
declared distribution refuses rather than quietly fitting a different model
than the one written -- `npag`/`npb` model the random effect distribution
nonparametrically, so a declared one contradicts them; `nlme` and `nls` are
Gaussian by construction; `vae`/`emvi`/`fbvi` hardcode the normal family in
their ELBO:

```{r refuse, error = TRUE}
nlmixr2(gammaCl, nlmixr2data::theo_sd, est = "npag")
```

`nlmixr2bayes`' `est = "nuts"` supports it too, and babelmixr2 translates the
families with an elementary quantile function to NONMEM.

## Starting values matter more than usual

These are harder likelihood surfaces than the models nlmixr2 usually sees, and
this is the practical thing to know about them.

Validated against Bauer's own gamma-distributed CL/V1 dataset (300 subjects):
at NONMEM's own estimates nlmixr2's objective function and per-subject
empirical Bayes estimates of the latent normals agree with NONMEM's to a
correlation of 0.999, and FOCEi warm started there converges near both the
simulation truth and NONMEM's answer. From a *cold* start, though,
`focei`/`laplace`/`saem` can settle into a worse local optimum -- one where the
residual error absorbs variability that belongs to the random effect.

This is the same thing Bauer reports ("the Laplace method does not travel well,
and tends to end prematurely when not near the answer"), and why his own recipe
runs an EM pass with `MCETA=10`-`100` before Laplace.

nlmixr2 has the same knob, and **a declared distribution turns it on by
default**: `mceta = 10` rather than the usual `-2`, announced when it happens.
`mceta = n` evaluates the previous eta, eta = 0, and `n - 1` draws from omega
for each subject's inner problem and keeps the best; for a declared random
effect the latent omega is the identity, so those draws are standard normal --
exactly the proposal `MCETA` uses.

This applies to the methods that have the knob: the FOCEi family,
`imp`/`impmap` and `agq`. `saem` has no `mceta` -- it *samples* the etas rather
than optimizing them, so there is no inner optimization to restart -- and its
control is left alone.

On that same Bauer dataset, cold started, it is worth what it costs:

| | truth | NONMEM (ITS) | `mceta = -2` | `mceta = 10` |
|---|---|---|---|---|
| gamma mean CL | 5.03 | 5.02 | 6.74 | 5.22 |
| gamma mean V1 | 4.66 | 4.69 | 8.02 | 5.55 |
| rel. variance CL | 0.085 | 0.085 | 0.225 | 0.145 |
| rel. variance V1 | 0.084 | 0.080 | 0.297 | 0.134 |

Better, but not the whole way -- the residual error is still overstated. So if
the fit looks like it stopped early, do what Bauer does: run an EM method
first (`est = "impmap"`/`"imp"`) and warm start `focei` from it, raise `mceta`
further, or give the model better starting values. Set `mceta` explicitly to
take charge of it. For `saem`, where the knob does not exist, the warm start
and the starting values are the levers you have.
