---
title: "Non-Gaussian random effect distributions"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Non-Gaussian random effect distributions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>",
                      message = FALSE, warning = FALSE)
## Every fit below is run for real and cached under inst/cache with
## nlmixr2save's `:=` operator, so this page shows live output without
## refitting on every build.  See vignettes/cacheSetup.R.
source("cacheSetup.R")
estCache("nonGaussianEta-")
```

A random effect in nlmixr2 is Normal by construction. Sometimes it should not
be: a heavy-tailed random effect is the standard robustness move against
influential individuals, and a clearance that is *itself* gamma distributed is
a different model from a log-normal one, not a reparameterization of it.

You declare the distribution in the `ini({})` block with a `dist()` line:

```{r model}
library(nlmixr2est)

gammaCl <- function() {
  ini({
    tka   <- 0.45
    lclm  <- log(1.1)   # gamma mean of CL
    lclrv <- log(0.09)  # gamma relative variance of CL
    tv    <- 3.45
    eta.ka ~ 0.6
    dist(eta.cl) ~ dgamma(shape = 1/exp(lclrv),
                          rate  = 1/(exp(lclrv)*exp(lclm)))
    add.sd <- 0.7
  })
  model({
    ka <- exp(tka + eta.ka)
    cl <- eta.cl        # a gamma distributed clearance
    v  <- exp(tv)
    d/dt(depot)  <- -ka*depot
    d/dt(center) <-  ka*depot - cl/v*center
    cp <- center/v
    cp ~ add(add.sd)
  })
}
```

Two things to notice. `eta.cl` **is** the clearance -- the declared
distribution is the marginal distribution of the random effect itself, so the
model block uses it directly rather than as a deviation from a typical value.
And it has no variance: the declaration already fixes its marginal, and the
latent scale it is built on is standard normal, so `dist()` declares the
random effect and implies `~ 1`. The reason that has to be one is below.

`lotri::lotriEtaDists()` lists the families you can declare:

```{r families}
head(lotri::lotriEtaDists()[, c("name", "parNames", "support")], 12)
```

## How it works

The implementation is Bauer's (NONMEM 7.5.1): **keep the latent random effect
standard normal and change the CDF**.

```
z   ~ N(0, 1)          latent, unit variance, fixed
u   = phiU(z)          normal CDF        ->  U(0, 1)
eta = Q(u; args)       inverse CDF of the declared family
```

`rxEtaDistExpand()` performs that rewrite, and it is worth looking at once,
because everything downstream -- simulation and every estimation method -- sees
only its output:

```{r expand}
rxode2::rxEtaDistExpand(gammaCl())$lstExpr[1:3]
```

The declared random effect became an ordinary model variable computed from a
latent standard normal, and the arguments of `dgamma()` are ordinary `ini({})`
parameters, estimated like any others. That is the whole trick: **the
estimation machinery never needs a non-Gaussian prior**, because the random
effect it integrates over is still standard normal.

### Why the variance has to be one

Correlation between two declared random effects is induced on the *latent*
scale, through a Cholesky factor -- a Gaussian copula. That makes the block a
**correlation** matrix rather than a covariance one, so its diagonal has to be
one. That is the one case where you write the block out, because the
correlation has nowhere else to go:

```r
eta.cl + eta.v ~ c(1,
                   0.5, 1)
dist(eta.cl) ~ dgamma(...)
dist(eta.v)  ~ dgamma(...)
```

`lotri()` says so if the diagonal is not one. nlmixr2 cannot fix individual
components of an omega block, so this is stated by *estimating* the diagonal at
one rather than by fixing it, and the expansion turns the off-diagonals into
unconstrained `rxCor.*` parameters with a fixed identity omega -- a form every
method already handles. The fit reports the correlation back:
`fit$etaDistCor`, and `tanh()` of an `rxCor.*` estimate (its back-transformed
column) is that correlation directly.

## Fitting

Nothing special at the call site:

```{r fit}
fit := nlmixr2(gammaCl, nlmixr2data::theo_sd, est = "focei",
               control = foceiControl(print = 0))
```

```{r print}
print(fit)
```

The gamma's own arguments come back as ordinary population parameters, and the
fit carries both the latent normal and the gamma-distributed clearance per
subject:

```{r etas}
head(fit$eta)
head(fit[, c("ID", "rxz.eta.cl", "eta.cl")], 3)
```

## Which methods support this

The FOCEi family (`focei`/`foce`/`laplace`/`agq`/`imp`/`impmap` and their
variants), `saem`, `posthoc` and simulation. A method that cannot use a declared
distribution refuses rather than quietly fitting a different model than the
one written -- `npag`/`npb` model the random effect distribution
nonparametrically, so a declared one contradicts them; `nlme` and `nls` are
Gaussian by construction; `vae`/`emvi`/`fbvi` hardcode the normal family in
their ELBO:

```{r refuse, error = TRUE}
nlmixr2(gammaCl, nlmixr2data::theo_sd, est = "npag")
```

`nlmixr2bayes`' `est = "nuts"` supports it too, and babelmixr2 translates the
families with an elementary quantile function to NONMEM.

## Getting these models to converge

These are harder likelihood surfaces than the models nlmixr2 usually sees, and
this is the practical thing to know about them.

Validated against Bauer's own gamma-distributed CL/V1 dataset (300 subjects):
at NONMEM's own estimates nlmixr2's objective function and per-subject
empirical Bayes estimates of the latent normals agree with NONMEM's to a
correlation of 0.999. The *likelihood* is not what differs. What differs is
how well the fit travels to it.

The reason is structural. `rxEtaDistExpand()` writes the declared
distribution's parameters into the model as bare thetas inside an inverse-CDF
call, and nothing about that shape is `theta + eta`. So every one of them
comes out **non**-mu-referenced -- which is the case both `saem` and the FOCEi
family handle worst. A cold start of either method alone settles a long way
from the answer.

### Starting values matter a great deal

The reason is structural. `rxEtaDistExpand()` writes the declared
distribution's parameters into the model as bare thetas inside an inverse-CDF
call, and nothing about that shape is `theta + eta`. So every one of them
comes out **non**-mu-referenced, which is the case the FOCEi family handles
worst. A cold start settles a long way from the answer.

What that costs, and what good starting values buy, on Bauer's data:

| | CL | Q | V2 | prop.sd | corr |
|---|---|---|---|---|---|
| simulation truth | 5.03 | 2.13 | 10.34 | 0.141 | 0.44 |
| `focei` from good starting values | 5.04 | 2.15 | 11.39 | 0.196 | 0.58 |

That fit reached an objective of -9099.8, against -8602.2 for the same
objective evaluated at NONMEM's own estimates. The structural parameters land
essentially on the truth; the variance components stay inflated.

Two cautions on reading that table. The correlation is
`tanh(rxCor.eta.v1.eta.cl)`, not the `rxCor.eta.v1.eta.cl` row `parFixed`
prints -- that row is the *unconstrained* copula parameter, and reading it as
a correlation overstates it. Ask the fit for `fit$etaDistCor` instead. And
the objective is still sensitive to the inner problem (below), so an
objective gap of a few hundred units between two parameter vectors is not yet
a reliable model comparison.

Getting to starting values that good currently takes work, because the EM
method you would normally warm-start from is unavailable here -- see
`saem` above. `rxode2::rxEtaDistMuRef()` will mu-reference the declared
distribution's parameters for you, which is the structure NONMEM uses
(`MU_5 = THETA(5)` with `$OMEGA (0.0 FIXED)`), but it does not make `saem`
sample the declared random effect, so it is not on its own a way around that
limitation.

### The inner problem

Even with good starting values the inner (eta) problem needs care, because the
inverse CDF is strongly non-linear and each subject's problem has real
curvature. Hold the parameters *fixed* and change only
`foceiControl(mceta=)`, which controls where each inner optimization starts,
and the objective moves by thousands of units at the same parameters. The
outer optimizer differentiates that objective, so with a poor inner solve it
is not searching a stable function -- which is how a fit can report
*"last objective function was not at minimum"*.

So fit with `foceiControl(mceta = 100)`, or at least `mceta = 0`.

`saem` has no `mceta` -- it *samples* the etas rather than optimizing them, so
there is no inner optimization to restart -- and there the recipe above and
the starting values are the levers you have.
