---
title: "How nlmixr2's SAEM differs from NONMEM's and saemix's"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{How nlmixr2's SAEM differs from NONMEM's and saemix's}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

Three widely used implementations of SAEM disagree in ways that are invisible
from the outside and that matter a great deal on hard models.  This page is a
line-by-line comparison of

* **NONMEM** `$EST METHOD=SAEM`, as specified in the *NONMEM 7 Technical Guide*
  (R.J. Bauer) -- the MCMC kernels in its section "The MCMC method of
  Expectation in SAEM" (eqs. 1.133-1.153) and the maximization in eqs.
  1.45-1.52;
* **saemix** 3.5 (Comets, Lavielle, Mentre), the reference R implementation of
  Kuhn & Lavielle's algorithm -- `R/main_estep.R`, `R/main_mstep.R`,
  `R/main_initialiseMainAlgo.R`;
* **nlmixr2est** `est = "saem"` -- `src/saem.cpp`, `R/saem_fit.R`,
  `R/saemControl.R`.

Everything below is sourced to one of those three.  Where nlmixr2 has an option
that closes a gap, it is named; where it does not, that is said plainly.

## The shape all three share

All three are the same algorithm:

1. **E-step.**  Draw each subject's parameter vector by Metropolis-Hastings
   from `p(phi_i | y_i, mu, Omega)`, starting from the previous iteration's
   draw (so it is a Markov chain across iterations, not a fresh chain).
2. **M-step.**  Update the population parameters from sufficient statistics
   accumulated with a Robbins-Monro gain, `s_k = s_{k-1} + gamma_k (s(phi_k) -
   s_{k-1})`.
3. **Two phases.**  A first phase with `gamma_k = 1` (full replacement, so the
   chain can travel), then a second with `gamma_k = 1/k` (averaging, so it
   converges).

The gain schedule is effectively identical in all three: `1` throughout the
burn-in and `1/k` afterwards (saemix `R/main_initialiseMainAlgo.R:282`,
nlmixr2 `R/saem_fit.R:812` with `mcmc$stepsize = 0:1`, NONMEM eqs. 1.146 and
1.152).  saemix additionally zeroes the gain for the first `nbiter.burn = 5`
iterations.

The differences are in the E-step's kernels, in what happens to a parameter
without between-subject variability, and in how far each implementation goes
in a single maximization.

## Difference 1: the MCMC kernels

| | NONMEM | saemix | nlmixr2 |
|---|---|---|---|
| independence from the prior `N(mu_i, Omega)` | mode 1, `ISAMPLE_M1` | kernel 1, `nbiter.mcmc[1] = 2` | kernel 1, `nu[1] = 2` |
| proposal from each subject's **accumulated conditional mean/variance** | **mode 1B**, from iteration 10 on | none | kernel 4, `nu1B` -- **default `0`, i.e. off** |
| borrow another subject's draw | mode 1A (off by default) | none | none |
| **multivariate** random walk | mode 2, proposal `Z = lambda * Omega` | -- | -- |
| **diagonal** block random walk | -- | kernel 2, `nbiter.mcmc[2] = 2` | kernel 2, `nu[2] = 2` |
| coordinate-wise random walk | mode 3, from `diag(Z^-1)` | kernel 3, `nbiter.mcmc[3] = 2` | kernel 3, `nu[3] = 2` |
| MAP-mode proposal | -- | kernel 4, `nbiter.mcmc[4] = 0` (off) | none |
| acceptance-rate adaptation | `IACCEPT`, scale tuned **per subject**, bounded by `ISCALE_MIN`/`MAX` | `proba.mcmc = 0.4`, one scale per coordinate, population-wide | `iaccept = 0.234` / `iacceptSingle = 0.44`, one scale per coordinate, population-wide |

Two of these are worth dwelling on.

**NONMEM's random walk is `Omega`-shaped; the other two walk on a diagonal.**
NONMEM's mode 2 proposes from `N(phi_i, lambda * Omega)` (eq. 1.139) -- the
full covariance, so the walk moves along the correlated directions of the
posterior.  saemix's kernel 2 (`domega2`, `R/main_estep.R:44`) and nlmixr2's
kernel 2 (`Gdiag_phi`, `src/saem.cpp` `do_mcmc` case 2) propose one coordinate
at a time with independent steps.  On a model whose random effects are strongly
correlated -- which is exactly what a copula, or a two-compartment CL/V pair,
produces -- a diagonal walk mixes slowly along the correlated direction while
the acceptance rate looks perfectly healthy.  Nothing reports this.

**NONMEM adapts the proposal scale per subject; the other two adapt it
globally.**  NONMEM tunes `lambda` for each subject so that subject's
acceptance rate approaches `IACCEPT` (eq. 1.139 and the surrounding text).
saemix and nlmixr2 tune one scale per coordinate against a pooled acceptance
rate (saemix `R/main_estep.R:59`, nlmixr2 `adaptRw()`).  A pooled rate of 0.3
is compatible with half the subjects at 0.6 and half at 0.0; the half that is
stuck never moves, and its draws are the previous iteration's draws.

nlmixr2 is the only one of the three that implements NONMEM's mode 1B, but it
is **off by default** (`saemControl(nu1B = 0)`).  Setting `nu1B = 2` with
`nb1B = 10` reproduces NONMEM's arrangement.

## Difference 2: a parameter with no between-subject variability

This is the largest structural difference and the one most likely to change an
answer.

The mu-referenced M-step is a generalized least squares normal equation
weighted by `Omega^-1` (NONMEM eqs. 1.45/1.46, saemix `R/main_mstep.R:63`,
nlmixr2 `src/saem.cpp`).  It cannot move a theta whose random effect has
(near-)zero variance: the column gets an enormous weight, the solve reduces to
"reproduce this column's sampled mean", and the sampler cannot move the column
off its prior mean -- which *is* the current theta.  The fixed point is the
`ini()` value.

All three implementations know this.  They solve it differently.

**NONMEM** derives a second route (eqs. 1.47-1.52) for thetas not expressed
through mu.  It differentiates the *entire* joint density, builds a
**per-subject outer-product (BHHH) information matrix** `H = sum_i g_i g_i'`,
and takes **one** damped Newton step `theta + H^-1 g` per iteration.  The
damping is an explicit **acceptance test**: try `alpha = 1`, evaluate the
complete-data objective, and if it did not improve, shrink `alpha` by
`sqrt(2)` and try again (the text after eq. 1.46).  The accepted step is then
averaged across iterations, `theta_Sk = ((k-1) theta_S,k-1 + theta_k)/k`
(eq. 1.152).

**saemix** drops such parameters out of the GLS
(`covariate.estim1[, i0.omega2] <- 0`) and hands them to
`optim(compute.Uy, maxit = 100)` -- a **full Nelder-Mead maximization** of the
data likelihood at the current draws -- then damps the result by the gain
(`R/main_mstep.R:66-71`).  It also does two things to keep that stable: the
direct route is **switched on only after the annealing phase**
(`kiter >= nbiter.sa`, half the burn-in), and until then such a parameter is
given a **real, decaying random effect**, shrunk every iteration by
`alpha0.sa = 10^(-3/nbiter.sa)` -- a 1000-fold decay across the phase
(`R/main_mstep.R:91`).

**nlmixr2** is closest to saemix: `refinePhi0Lik()` runs a full bounded
derivative-free maximization (bobyqa / newuoa / Nelder-Mead, budget
`nonMuThetaMaxEval`) and damps the argmax by the gain.  The equivalents of
saemix's two stabilizers exist but are **not the default**:

| | saemix | nlmixr2 |
|---|---|---|
| exploration width for a zero-variance random effect | `omega.init`, decayed | `zeroOmegaTune = 0.1`, **held constant** |
| decay that width across annealing | always, 1000-fold | `zeroOmegaAnneal = 0` (off); set `1000` to match |
| direct maximization instead of the GLS | always, after `nbiter.sa` | `zeroOmegaDirect = FALSE` (off) |
| when the direct route starts | half the **burn-in** | `round((nBurn + nEm)/2)` -- half the **total** -- or `nonMuThetaStart` |

The last row is a real discrepancy rather than a preference.  With the package
defaults (`nBurn = 200`, `nEm = 300`) nlmixr2 does not start refining a non-mu
theta until iteration 250 of 500, where the gain is already `~1/50`; saemix's
equivalent starts at iteration 150 of 400, while the gain is still `1`.
`saemControl(nonMuThetaStart =)` sets it directly.

Two further differences inside the step itself:

* nlmixr2 accumulates its score and information **per observation**
  (`src/nonMuThetaGrad.h`), NONMEM **per subject** (eq. 1.51).  The
  per-observation outer product ignores the within-subject correlation of the
  score and is not the same matrix.
* Neither nlmixr2 nor saemix applies NONMEM's `alpha` **acceptance test**.
  Both damp an unconditional argmax.  A full maximization of a nearly flat or
  degenerate direction goes to the boundary, and damping only controls how fast
  it marches there; NONMEM's accepted single step cannot.

## Difference 3: simulated annealing

saemix anneals during the first `nbiter.sa = nbiter.saemix[1]/2` iterations:
the diagonal of `Omega` is floored so it cannot fall faster than
`alpha.sa = 0.97` per iteration, and the residual error is floored the same way
(`R/main_mstep.R:85-95, 100-115`).  This is what stops a variance from
collapsing early, before the chain has explored.

nlmixr2 has an annealing phase (`perSa`, `coefSa = 0.95`) and NONMEM has
`CONSTRAIN`/`CTYPE`.  What nlmixr2 does **not** have by default is saemix's
decay of the exploration width for parameters without between-subject
variability -- see `zeroOmegaAnneal` above.

## Difference 4: chains

| | NONMEM | saemix | nlmixr2 |
|---|---|---|---|
| chains per iteration | `ISAMPLE`, default 2 for SAEM | `nb.chains = 1` | `nmc = 3` |

## What none of them has: an M-step on the random effect distribution

`nlmixr2est` can declare a non-Gaussian random effect distribution (see
`vignette("nonGaussianEta")`).  It is worth being explicit that **neither
NONMEM nor saemix has an M-step for such a distribution's parameters**, because
neither needs one:

* saemix has no non-Gaussian random effect families at all.  `transform.par`
  gives log/logit/probit *scales*, but the random effect is Normal on that
  scale.
* NONMEM's technique (Bauer, `gamma_indpar`) keeps the random effect a
  standard normal and moves the family into `$PK` through an inverse CDF.  The
  family's parameters are then **ordinary thetas**, updated by whichever of the
  two routes above applies.

nlmixr2's `saemControl(etaDistMstep = TRUE)` is an addition with no counterpart
in either: it refits the family directly to the sampled random effects,
skipping the ODE entirely.  In the `(y, eta)` augmentation the complete-data
likelihood factors as `log p(y | eta) + log p(eta | theta_dist)`, and
`theta_dist` appears only in the second term, so this is a valid M-step -- but
only if it is fed genuine, well-mixed posterior **draws**.  The failure modes
are worth showing, because they are silent:

```{r mstepDemo}
## eta ~ Gamma(mean 5.03, relative variance 0.086), written Bauer's way as
## eta = qgamma(pnorm(z)) with z ~ N(0,1); one noisy observation per subject.
set.seed(42)
N <- 300; sdE <- 1; mu0 <- 5.03; rv0 <- 0.086
sh0 <- 1/rv0; ra0 <- 1/(rv0*mu0)
y <- qgamma(pnorm(rnorm(N)), sh0, ra0) + rnorm(N, 0, sdE)

Q   <- function(z, sh, ra) qgamma(pnorm(z), sh, ra)
llY <- function(z, sh, ra) dnorm(y, Q(z, sh, ra), sdE, log = TRUE)

## E-step: Metropolis draws of the latent normal
estep <- function(z, sh, ra, nmh, step) {
  cur <- llY(z, sh, ra) + dnorm(z, log = TRUE)
  for (k in seq_len(nmh)) {
    zc <- z + rnorm(N, 0, step)
    pc <- llY(zc, sh, ra) + dnorm(zc, log = TRUE)
    ok <- (pc - cur) > log(runif(N)); z[ok] <- zc[ok]; cur[ok] <- pc[ok]
  }
  z
}
## M-step: refit the family to the IMPLIED random effects Q(pnorm(z); theta_old)
fitGamma <- function(e) {
  exp(optim(c(log(mean(e)^2/var(e)), log(mean(e)/var(e))),
            function(p) -sum(dgamma(e, exp(p[1]), exp(p[2]), log = TRUE)))$par)
}
run <- function(nmh, step, means = FALSE, niter = 200) {
  sh <- sh0; ra <- ra0; z <- rnorm(N)
  for (it in seq_len(niter)) {
    z  <- estep(z, sh, ra, nmh, step)
    zi <- if (means) rowMeans(replicate(20, estep(z, sh, ra, 3, step))) else z
    p  <- fitGamma(Q(zi, sh, ra))
    g  <- if (it > niter/2) 1/(it - niter/2) else 1   # the SAEM gain
    sh <- sh + g*(p[1] - sh); ra <- ra + g*(p[2] - ra)
  }
  c(mean = sh/ra, relVar = 1/sh)
}
rbind(`well-mixed draws`      = run(30, 0.6),
      `draws, barely mixing`  = run(1,  0.05),
      `conditional MEANS`     = run(30, 0.6, means = TRUE),
      truth                   = c(mu0, rv0))
```

Fed well-mixed draws the step sits on the truth.  Fed conditional means
instead of draws, the relative variance collapses to nothing while the mean
still looks right -- the most dangerous of the three, because the fit reports a
plausible typical value with no between-subject variability left.  Fed a chain
that has not mixed, the mean goes too.  Both fail *quietly*, looking like
convergence.  This is why `etaDistMstep` defaults to `FALSE`, why it carries a
latent-spread guard (`etaDistSdLo`/`etaDistSdHi`), and why the guard cannot be
removed.

Note also what the guard cannot see: for a copula member the quantity checked
is the *combined* latent `rho*z_j + sqrt(1-rho^2)*z_k`, whose spread is 1 by
construction for **any** `rho`, including a degenerate `rho = 1`.  A spread
check cannot detect copula collapse.

## Transcribing Bauer's own control stream

Bauer's `gamma_indpar` SAEM stream is the reference answer for the declared
gamma model, and it is worth reading for what it does *not* do:

```
$OMEGA BLOCK(2) VALUES(0.3,0.01)   ; ETA(1),ETA(2)  Q and V2, real IIV
$OMEGA (1.0 FIXED)X2               ; ETA(3),ETA(4)  the two latent normals
$OMEGA (0.0 FIXED)X5               ; ETA(5)..ETA(9) MU_ helper etas
```

with `MU_5..MU_9` carrying the two gamma means, the two relative variances and
the copula element `L21`.  So:

* the latent normals are **diagonal** unit-variance; the correlation lives in
  `$PK` as `L21*ETA(3) + sqrt(1-L21^2)*ETA(4)`, not in `$OMEGA` -- which is
  what `rxEtaDistExpand()` reproduces;
* **every** distribution parameter, and the copula element, is
  **mu-referenced with a helper random effect of variance exactly zero**.  None
  of them is a non-mu theta.

The nlmixr2 equivalent of the third line is `rxode2::rxEtaDistMuRef()`, and the
exact-zero variance needs the direct route:

```r
ui <- rxEtaDistMuRef(ui, variance = 0)
nlmixr2(ui, data, est = "saem",
        control = saemControl(zeroOmegaDirect = TRUE, zeroOmegaAnneal = 1000))
```

A non-zero `zeroOmegaTune` makes each helper a *real* random effect on the
gamma mean, the relative variance and `rho`, so each subject gets its own --
which absorbs between-subject variability that belongs to the latent normals
and leaves them under-dispersed.  That is a different model from the one
written, and it fits better in-sample.

Bauer's stream also runs `METHOD=ITS ... MCETA=100` for 100 iterations
*before* SAEM, and gives SAEM `NITER=500` with `IACCEPT=0.3`.  A cold SAEM
start at 200 iterations is not the same experiment.

## Summary of levers

| Gap | nlmixr2 setting that closes it | Default |
|---|---|---|
| no mode 1B kernel | `saemControl(nu1B = 2, nb1B = 10)` | off |
| GLS cannot move a zero-variance mu theta | `zeroOmegaDirect = TRUE` | `FALSE` |
| exploration width never decays | `zeroOmegaAnneal = 1000` | `0` |
| non-mu refinement starts at half the *total* | `nonMuThetaStart = round(nBurn/2)` | half of `nBurn + nEm` |
| single chain vs several | `nmc` | `3` |

Not closable by an option, and listed so they are not mistaken for tuning:
the diagonal (rather than `Omega`-shaped) random walk, the population-wide
(rather than per-subject) acceptance adaptation, the per-observation (rather
than per-subject) BHHH information, and the absence of NONMEM's `alpha`
acceptance test on the non-mu step.

## References

* Bauer, R.J. *NONMEM 7 Technical Guide*, sections on the MCMC method of
  Expectation in SAEM (eqs. 1.133-1.153) and the maximization step
  (eqs. 1.45-1.52).
* Bauer, R.J. NONMEM Tutorial Part II: Estimation Methods and Advanced Examples.
  *CPT Pharmacometrics Syst Pharmacol* 2019.
* Comets, E., Lavielle, M., Mentre, F. Parameter estimation in nonlinear mixed
  effect models using saemix. *J Stat Softw* 2017; and the saemix 3.5 sources.
* Kuhn, E., Lavielle, M. Coupling a stochastic approximation version of EM with
  an MCMC procedure. *ESAIM P&S* 2004.
