---
title: "How nlmixr2's SAEM differs from NONMEM's, Monolix's and saemix's"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{How nlmixr2's SAEM differs from NONMEM's, Monolix's and saemix's}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

Three widely used implementations of SAEM disagree in ways that are invisible
from the outside and that matter a great deal on hard models.  This page is a
line-by-line comparison of

* **NONMEM** `$EST METHOD=SAEM`, as specified in the *NONMEM 7 Technical Guide*
  (R.J. Bauer) -- the MCMC kernels in its section "The MCMC method of
  Expectation in SAEM" (eqs. 1.133-1.153) and the maximization in eqs.
  1.45-1.52;
* **Monolix**, as specified in its own *Methodology* document (sections 1.1.1
  "The SAEM algorithm", 1.1.2 "The MCMC-SAEM algorithm" and 1.1.3 "The
  Simulated Annealing SAEM algorithm", eqs. 1.1 and 1.12-1.15);
* **saemix** 3.5 (Comets, Lavielle, Mentre), the reference R implementation of
  Kuhn & Lavielle's algorithm -- `R/main_estep.R`, `R/main_mstep.R`,
  `R/main_initialiseMainAlgo.R`;
* **nlmixr2est** `est = "saem"` -- `src/saem.cpp`, `R/saem_fit.R`,
  `R/saemControl.R`.

Everything below is sourced to one of those four.  Where nlmixr2 has an option
that closes a gap, it is named; where it does not, that is said plainly.

Monolix and saemix are the same algorithm by descent -- both are Lavielle's --
so they agree on most of what follows.  Where they *disagree* is unusually
informative, because it separates "this is the Kuhn-Lavielle algorithm" from
"this is one implementation's choice".

## The shape all three share

All three are the same algorithm:

1. **E-step.**  Draw each subject's parameter vector by Metropolis-Hastings
   from `p(phi_i | y_i, mu, Omega)`, starting from the previous iteration's
   draw (so it is a Markov chain across iterations, not a fresh chain).
2. **M-step.**  Update the population parameters from sufficient statistics
   accumulated with a Robbins-Monro gain, `s_k = s_{k-1} + gamma_k (s(phi_k) -
   s_{k-1})`.
3. **Two phases.**  A first phase with `gamma_k = 1` (full replacement, so the
   chain can travel), then a second with `gamma_k = 1/k` (averaging, so it
   converges).

The gain schedule is effectively identical in all three: `1` throughout the
burn-in and `1/k` afterwards (saemix `R/main_initialiseMainAlgo.R:282`,
nlmixr2 `R/saem_fit.R:812` with `mcmc$stepsize = 0:1`, NONMEM eqs. 1.146 and
1.152).  saemix additionally zeroes the gain for the first `nbiter.burn = 5`
iterations.

The differences are in the E-step's kernels, in what happens to a parameter
without between-subject variability, and in how far each implementation goes
in a single maximization.

## Difference 1: the MCMC kernels

| | NONMEM | Monolix | saemix | nlmixr2 |
|---|---|---|---|---|
| independence from the prior `N(C_i mu, Omega)` | mode 1, `ISAMPLE_M1` | kernel 1 | kernel 1, `nbiter.mcmc[1] = 2` | kernel 1, `nu[1] = 2` |
| proposal from each subject's **accumulated conditional mean/variance** | **mode 1B**, from iteration 10 on | none | none | kernel 4, `nu1B` -- **default `0`, i.e. off** |
| borrow another subject's draw | mode 1A (off by default) | kernel 2 (**deprecated**) | none | none |
| **multivariate** random walk, `Omega`-shaped | mode 2, `Z = lambda * Omega` | kernel 4, `N(phi, kappa * Omega)`, block size cycling 2..d | -- | -- |
| **diagonal** block random walk | -- | -- | kernel 2, `nbiter.mcmc[2] = 2`; kernel 3, block size cycling | kernel 2, `nu[2] = 2`; kernel 3 |
| coordinate-wise random walk | mode 3, from `diag(Z^-1)` | kernel 3 | kernel 3, `nbiter.mcmc[3] = 2` | kernel 3, `nu[3] = 2` |
| MAP-mode proposal | -- | -- | kernel 4, `nbiter.mcmc[4] = 0` (off) | none |
| acceptance-rate adaptation | `IACCEPT`, scale tuned **per subject**, bounded by `ISCALE_MIN`/`MAX` | `kappa` | `proba.mcmc = 0.4`, one scale per coordinate, population-wide | `iaccept = 0.234` / `iacceptSingle = 0.44`, one scale per coordinate, population-wide |

Two of these are worth dwelling on.

**NONMEM's and Monolix's random walks are `Omega`-shaped; saemix's and
nlmixr2's walk on a diagonal.**  NONMEM's mode 2 proposes from
`N(phi_i, lambda * Omega)` (eq. 1.139) and Monolix's kernel 4 from
`N(phi_i, kappa * Omega)` -- the full covariance, so the walk moves along the
correlated directions of the posterior.  saemix's kernel 2 (`domega2`,
`R/main_estep.R:44`) and nlmixr2's kernel 2 (`Gdiag_phi`, `src/saem.cpp`
`do_mcmc` case 2) propose with independent per-coordinate steps.  On a model
whose random effects are strongly correlated -- which is exactly what a copula,
or a two-compartment CL/V pair, produces -- a diagonal walk mixes slowly along
the correlated direction while the acceptance rate looks perfectly healthy.
Nothing reports this.

This is the clearest case where Monolix and saemix part company, and it is the
tell: saemix's kernel 3 reproduces Monolix's kernel 4 *structure* (blocks of
consecutive components, block size cycling from 2 up to `d`) but takes a
**diagonal** step where Monolix takes `kappa * Omega`.  nlmixr2 inherited
saemix's version.  `saemControl(rwOmega = TRUE)` switches nlmixr2's kernel 2 to
the `Omega`-shaped proposal.

**NONMEM adapts the proposal scale per subject; the others adapt it
globally.**  NONMEM tunes `lambda` for each subject so that subject's
acceptance rate approaches `IACCEPT` (eq. 1.139 and the surrounding text).
saemix and nlmixr2 tune one scale per coordinate against a pooled acceptance
rate (saemix `R/main_estep.R:59`, nlmixr2 `adaptRw()`).  A pooled rate of 0.3
is compatible with half the subjects at 0.6 and half at 0.0; the half that is
stuck never moves, and its draws are the previous iteration's draws -- which is
precisely the input that makes a distribution M-step collapse (see below).
`saemControl(iacceptPerId = TRUE)` gives nlmixr2 one scale per subject.

nlmixr2 is the only one of the three that implements NONMEM's mode 1B, but it
is **off by default** (`saemControl(nu1B = 0)`).  Setting `nu1B = 2` with
`nb1B = 10` reproduces NONMEM's arrangement.

## Difference 2: a parameter with no between-subject variability

This is the largest structural difference and the one most likely to change an
answer.

The mu-referenced M-step is a generalized least squares normal equation
weighted by `Omega^-1` (NONMEM eqs. 1.45/1.46 and 1.148-1.150, saemix
`R/main_mstep.R:63`, nlmixr2 `src/saem.cpp`).  It is natural to say that this
update "cannot move a theta whose random effect has zero variance because the
weight explodes".  **That is not right, and the distinction matters**, so it is
worth doing the algebra once.

For an intercept-only mu on a coordinate `k` whose random effect is
uncorrelated with the rest, `d(mu_i)/d(theta) = 1` and

```
g = sum_i (1/w^2) (phibar_ik - theta)          H = sum_i (1/w^2) = m / w^2
H^-1 g = mean_i(phibar_ik - theta)
```

The `w^2` **cancels**.  The mu M-step is `theta <- mean_i(phibar_ik)` with no
variance in it anywhere, and it is perfectly well defined in the limit
`w -> 0`.  Nothing about the maximization degenerates.

What degenerates is the **E-step**.  If the random effect genuinely has zero
variance, the sampler cannot move it, so `phibar_ik = mu_ik = theta` for every
subject, the increment is identically zero, and the fixed point is the `ini()`
value.  The theta does not move because there is no variability to inform it --
not because the weighting broke.

That reading is confirmed from the other side inside this package.  `imp`'s
`impMuInterceptStep()` is literally `theta += mean(eta)` -- the same update --
and it works, because `imp` takes its etas from the inner MAP and drops a flat
random effect from `Omega^-1` entirely, so the MAP moves it freely against the
data and `mean(eta)` carries information.  `saem` cannot do that: it *samples*
the eta from a prior of variance ~0.

So a "second M-step route" is not the fix for a zero-variance mu-referenced
theta in `saem`, and in this package it has been measured not to be: forcing
the width non-zero so the sampler can move (`zeroOmegaTune`) degrades every
parameter monotonically on Bauer's gamma model (residual SD 0.150 / 0.162 /
0.170 at 0.1 / 1 / 4, against a truth of 0.141), and annealing that width
without a second route pins the theta outright (residual SD 0.72 -> 1.20 on the
theophylline helper fixture).  Put a parameter that has no between-subject
variability on the **non-mu** route instead, where the data likelihood, not a
sampled mean, is what moves it.

The rest of this section is about that non-mu route, which all four
implementations do have and implement differently.

**NONMEM** derives a second route (eqs. 1.47-1.52) for thetas not expressed
through mu.  It differentiates the *entire* joint density, builds a
**per-subject outer-product (BHHH) information matrix** `H = sum_i g_i g_i'`,
and takes **one** damped Newton step `theta + H^-1 g` per iteration.  The
damping is an explicit **acceptance test**: try `alpha = 1`, evaluate the
complete-data objective, and if it did not improve, shrink `alpha` by
`sqrt(2)` and try again (the text after eq. 1.46).  The accepted step is then
averaged across iterations, `theta_Sk = ((k-1) theta_S,k-1 + theta_k)/k`
(eq. 1.152).

**saemix** drops such parameters out of the GLS
(`covariate.estim1[, i0.omega2] <- 0`) and hands them to
`optim(compute.Uy, maxit = 100)` -- a **full Nelder-Mead maximization** of the
data likelihood at the current draws -- then damps the result by the gain
(`R/main_mstep.R:66-71`).  It also does two things to keep that stable: the
direct route is **switched on only after the annealing phase**
(`kiter >= nbiter.sa`, half the burn-in), and until then such a parameter is
given a **real, decaying random effect**, shrunk every iteration by
`alpha0.sa = 10^(-3/nbiter.sa)` -- a 1000-fold decay across the phase
(`R/main_mstep.R:91`).

**Monolix** does something different from all of them, and it is the cleanest
statement of the issue.  Its stochastic approximation is applied to the
**objective**, not to the parameter (eq. 1.1):

```
Q_k(theta) = Q_{k-1}(theta) + gamma_k ( log p(y, psi^(k); theta) - Q_{k-1}(theta) )
theta_{k+1} = argmax_theta Q_k(theta)
```

It averages the *function* over iterations and then maximizes; saemix and
nlmixr2 maximize the current iteration's function and then average the
*maximizer*.  For an exponential family these coincide -- which is why the
sufficient-statistic shortcut is valid for the mu thetas -- but for a non-mu
theta they do not.  Averaging first is much the more stable of the two on a
flat or degenerate direction: `Q_k` carries curvature accumulated from every
past draw, so no single draw's flat direction can send the argmax to a
boundary.  Maximizing first has nothing holding it back, and the gain then
only sets how fast the parameter marches there.

**nlmixr2** is closest to saemix: `refinePhi0Lik()` runs a full bounded
derivative-free maximization (bobyqa / newuoa / Nelder-Mead, budget
`nonMuThetaMaxEval`) and damps the argmax by the gain.
`saemControl(nonMuThetaBhhh = TRUE)` replaces that with NONMEM's arrangement --
one per-subject BHHH Newton step, subject to the `alpha` acceptance test, and
no search at all.  The equivalents of
saemix's two stabilizers exist but are **not the default**:

| | saemix | nlmixr2 |
|---|---|---|
| exploration width for a zero-variance random effect | `omega.init`, decayed | `zeroOmegaTune = 0.1`, **held constant** |
| decay that width across annealing | always, 1000-fold | `zeroOmegaAnneal = 0` (off); set `1000` to match |
| direct maximization instead of the GLS | always, after `nbiter.sa` | `zeroOmegaDirect = FALSE` (off) |
| when the direct route starts | half the **burn-in** | `round((nBurn + nEm)/2)` -- half the **total** -- or `nonMuThetaStart` |

The last row is a real discrepancy rather than a preference.  With the package
defaults (`nBurn = 200`, `nEm = 300`) nlmixr2 does not start refining a non-mu
theta until iteration 250 of 500, where the gain is already `~1/50`; saemix's
equivalent starts at iteration 150 of 400, while the gain is still `1`.
`saemControl(nonMuThetaStart =)` sets it directly.

Two further differences inside the step itself:

* nlmixr2 accumulates its score and information **per observation**
  (`src/nonMuThetaGrad.h`), NONMEM **per subject** (eq. 1.51).  The
  per-observation outer product ignores the within-subject correlation of the
  score and is not the same matrix.
* Neither nlmixr2 nor saemix applies NONMEM's `alpha` **acceptance test**.
  Both damp an unconditional argmax.  A full maximization of a nearly flat or
  degenerate direction goes to the boundary, and damping only controls how fast
  it marches there; NONMEM's accepted single step cannot, and Monolix's
  averaged objective cannot either.  nlmixr2's `nonMuThetaBhhh = TRUE` gates
  on the same test: try `alpha = 1`, evaluate `phi0Objective`, shrink by
  `sqrt(2)` up to eight times, and if nothing improves leave the theta alone.
  Note that this is a *different* gate from the default path's, which tests
  the **conditioning** of the information matrix (`nonMuGradStep()`) -- that
  says the step is numerically trustworthy, not that it goes downhill.

## Difference 3: simulated annealing

Monolix states this one most clearly (section 1.1.3).  It introduces a
**temperature** `T_k` and runs MCMC-SAEM against the tempered complete model
`p_T(y, phi; theta) proportional to exp(-U(y, phi; theta)/T)`, which "consists
in replacing the variance matrix `Omega` by `T Omega` and the residual variance
`a^2` by `T a^2`".  `T_k` starts large and decreases to 1.  In practice that is
implemented as a floor on how fast the variances may fall (eqs. 1.14/1.15):

```
Omega_{k+1} = max(tau * Omega_k, Omega~_{k+1})
a^2_{k+1}   = max(tau * a^2_k,   s3_k / N)
```

which is exactly saemix's `alpha.sa` rule -- saemix implements Monolix's
temperature.  Monolix additionally allows **two** coefficients, `tau1` for
`Omega` and `tau2` for `a^2`, and explicitly sanctions `tau1 < 1` with
`tau2 > 1`: fit tightly at first while allowing large between-subject
variability, then trade the two off.  Neither saemix nor nlmixr2 exposes the
second coefficient.

saemix anneals during the first `nbiter.sa = nbiter.saemix[1]/2` iterations:
the diagonal of `Omega` is floored so it cannot fall faster than
`alpha.sa = 0.97` per iteration, and the residual error is floored the same way
(`R/main_mstep.R:85-95, 100-115`).  This is what stops a variance from
collapsing early, before the chain has explored.

nlmixr2 has an annealing phase (`perSa`, `coefSa = 0.95`) and NONMEM has
`CONSTRAIN`/`CTYPE`.  What nlmixr2 does **not** have by default is saemix's
decay of the exploration width for parameters without between-subject
variability -- see `zeroOmegaAnneal` above.

## Difference 4: chains

| | NONMEM | Monolix | saemix | nlmixr2 |
|---|---|---|---|---|
| chains per iteration | `ISAMPLE`, default 2 for SAEM | `L`, recommended above 1 when `N` is small | `nb.chains = 1` | `nmc = 3` |
| iterations with the parameters frozen | `NBURN` | `K_b` ("we only run the MCMC algorithm but the parameters are not updated") | `nbiter.burn = 5` | -- |

That last row is worth noting on its own: NONMEM, Monolix and saemix all have a
phase in which the chain runs but nothing is estimated, and nlmixr2 has no
equivalent -- its `nBurn` is the full-replacement (`gamma = 1`) phase, which is
a different thing.  In Bauer's own gamma stream that phase is `NBURN=5000`
against `NITER=500`.

## What none of them has: an M-step on the random effect distribution

`nlmixr2est` can declare a non-Gaussian random effect distribution (see
`vignette("nonGaussianEta")`).  It is worth being explicit that **neither
NONMEM nor saemix has an M-step for such a distribution's parameters**, because
neither needs one:

* saemix has no non-Gaussian random effect families at all.  `transform.par`
  gives log/logit/probit *scales*, but the random effect is Normal on that
  scale.
* NONMEM's technique (Bauer, `gamma_indpar`) keeps the random effect a
  standard normal and moves the family into `$PK` through an inverse CDF.  The
  family's parameters are then **ordinary thetas**, updated by whichever of the
  two routes above applies.

nlmixr2's `saemControl(etaDistMstep = TRUE)` is an addition with no counterpart
in either: it refits the family directly to the sampled random effects,
skipping the ODE entirely.  In the `(y, eta)` augmentation the complete-data
likelihood factors as `log p(y | eta) + log p(eta | theta_dist)`, and
`theta_dist` appears only in the second term, so this is a valid M-step -- but
only if it is fed genuine, well-mixed posterior **draws**.  The failure modes
are worth showing, because they are silent:

```{r mstepDemo}
## eta ~ Gamma(mean 5.03, relative variance 0.086), written Bauer's way as
## eta = qgamma(pnorm(z)) with z ~ N(0,1); one noisy observation per subject.
set.seed(42)
N <- 300; sdE <- 1; mu0 <- 5.03; rv0 <- 0.086
sh0 <- 1/rv0; ra0 <- 1/(rv0*mu0)
y <- qgamma(pnorm(rnorm(N)), sh0, ra0) + rnorm(N, 0, sdE)

Q   <- function(z, sh, ra) qgamma(pnorm(z), sh, ra)
llY <- function(z, sh, ra) dnorm(y, Q(z, sh, ra), sdE, log = TRUE)

## E-step: Metropolis draws of the latent normal
estep <- function(z, sh, ra, nmh, step) {
  cur <- llY(z, sh, ra) + dnorm(z, log = TRUE)
  for (k in seq_len(nmh)) {
    zc <- z + rnorm(N, 0, step)
    pc <- llY(zc, sh, ra) + dnorm(zc, log = TRUE)
    ok <- (pc - cur) > log(runif(N)); z[ok] <- zc[ok]; cur[ok] <- pc[ok]
  }
  z
}
## M-step: refit the family to the IMPLIED random effects Q(pnorm(z); theta_old)
fitGamma <- function(e) {
  exp(optim(c(log(mean(e)^2/var(e)), log(mean(e)/var(e))),
            function(p) -sum(dgamma(e, exp(p[1]), exp(p[2]), log = TRUE)))$par)
}
run <- function(nmh, step, means = FALSE, niter = 200) {
  sh <- sh0; ra <- ra0; z <- rnorm(N)
  for (it in seq_len(niter)) {
    z  <- estep(z, sh, ra, nmh, step)
    zi <- if (means) rowMeans(replicate(20, estep(z, sh, ra, 3, step))) else z
    p  <- fitGamma(Q(zi, sh, ra))
    g  <- if (it > niter/2) 1/(it - niter/2) else 1   # the SAEM gain
    sh <- sh + g*(p[1] - sh); ra <- ra + g*(p[2] - ra)
  }
  c(mean = sh/ra, relVar = 1/sh)
}
rbind(`well-mixed draws`      = run(30, 0.6),
      `draws, barely mixing`  = run(1,  0.05),
      `conditional MEANS`     = run(30, 0.6, means = TRUE),
      truth                   = c(mu0, rv0))
```

Fed well-mixed draws the step sits on the truth.  Fed conditional means
instead of draws, the relative variance collapses to nothing while the mean
still looks right -- the most dangerous of the three, because the fit reports a
plausible typical value with no between-subject variability left.  Fed a chain
that has not mixed, the mean goes too.  Both fail *quietly*, looking like
convergence.  This is why `etaDistMstep` defaults to `FALSE`, why it carries a
latent-spread guard (`etaDistSdLo`/`etaDistSdHi`), and why the guard cannot be
removed.

Note also what the guard cannot see: for a copula member the quantity checked
is the *combined* latent `rho*z_j + sqrt(1-rho^2)*z_k`, whose spread is 1 by
construction for **any** `rho`, including a degenerate `rho = 1`.  A spread
check cannot detect copula collapse.

Measured on Bauer's gamma model (300 subjects, `nBurn = nEm = 100`, cold
start), the second failure mode is what actually happens:

| seed | `etaDistMstep` | CL | V1 | rel.var CL | rel.var V1 | cor |
|---|---|---|---|---|---|---|
| 7 | `FALSE` | 3.041 | 6.893 | 0.0469 | 0.1095 | 0.999 |
| 7 | `TRUE` | 0.515 | 0.105 | **0.0003** | **0.0027** | 0.999 |
| 99 | `FALSE` | 4.978 | 4.797 | 0.0521 | 0.0329 | 0.974 |
| 99 | `TRUE` | 3.621 | 4.066 | **0.0000** | **0.0000** | 0.996 |
| truth | | 5.03 | 4.66 | 0.086 | 0.086 | 0.438 |

Both relative variances collapse to zero, in both seeds -- the "conditional
means" row of the demonstration above, not the "well-mixed draws" row.  The
step is not wrong in principle; the chain it is being fed is not mixing well
enough to feed it.  This is why `etaDistMstep` defaults to `FALSE`, and why
the MCMC differences in Difference 1 are not academic.

### The M-step needs a mixing time between updates

The collapse above is at `etaDistEvery = 1` -- the default -- which runs the
M-step on **every** iteration.  That is the worst possible schedule for a step
whose only failure mode is being fed draws that have not moved, and spacing it
is what makes the step usable at all: on the repaired baseline
`etaDistEvery = 20` takes this model from 42.0% mean absolute relative error to
26.6% on its own, and to 10.0% alongside `iacceptPerId` (see the table above).

Two things are worth knowing before treating 20 as a default.  It is **one seed
and one model**, and it is a **fixed count** rather than a fraction of the run:
20 out of 200 iterations is 10% here, but the package default is 500 iterations
and a short fit might see the M-step fire once.  Whether the right rule is a
count or a fraction of `nBurn` is not established.

Note also that an earlier version of this page reported a much larger effect for
spacing alone (15.1%).  That was measured before the acceptance-adaptation
defects were fixed, when the schedule was partly compensating for a frozen
chain; on a healthy chain the honest figure is 26.6%, and the useful result is
the combination rather than the schedule by itself.

The copula M-step (`etaDistCorMstep`, which defaults **`TRUE`** and is keyed
off its own flag, so it runs even when `etaDistMstep = FALSE`) contributes
separately, and is gated by the same `etaDistEvery` -- so the default schedule
affects fits that never opt into `etaDistMstep` at all.  Same model and seed 99, with both off:

| | CL | V1 | rel.var CL | rel.var V1 | cor | MARE |
|---|---|---|---|---|---|---|
| `etaDistCorMstep = TRUE` (default) | 4.978 | 4.797 | 0.0521 | 0.0329 | 0.974 | 45.5% |
| `etaDistCorMstep = FALSE` | 5.020 | 4.872 | 0.0492 | 0.0336 | **0.729** | **35.0%** |
| truth | 5.03 | 4.66 | 0.086 | 0.086 | 0.438 | |

Its update is `rho <- cor(w_j, w_k)`, and `w_k` already contains `rho`, so the
recursion is `rho_new = rho + sqrt(1-rho^2) * c` where `c` is whatever
correlation the posterior draws have left over.  In isolation that is biased
upward but stable (a toy with `rho_true = 0.438` settles at ~0.60); on the real
model it accounts for roughly half of the remaining error in `rho`.

## Diagnosing the chain

Every difference in Difference 1 is a claim about how well the MCMC mixes, and
until recently nothing in `saem` could confirm or refute any of them: the
acceptance rate was computed each iteration, used to adapt the random-walk
scale, and discarded.  Four per-iteration traces now reach the fit.

```r
fit$env$mcmcAccept   # pooled acceptance, columns: prior, rw, coord, mode1B
fit$env$mcmcStuck    # fraction of SUBJECTS that accepted nothing
fit$env$mcmcPhiSd    # pooled SD of each sampled parameter
fit$env$mcmcPhiAcf   # lag-1 autocorrelation against the previous iteration
```

Two of these deserve emphasis.

`mcmcStuck` is the number a pooled acceptance rate structurally cannot show.
A population rate of 0.3 is equally consistent with every subject at 0.3 and
with 60% of subjects at 0.5 and 40% never moving at all.  Only the second is a
problem, and only the second contributes last iteration's draw to this
iteration's M-step.  This is also the diagnostic that says whether
`iacceptPerId = TRUE` is worth turning on.

`mcmcPhiSd` has a known target **for a declared non-Gaussian random effect**,
because that model pins down what the answer should be: the latent is standard
normal by construction with a fixed identity omega, and well-mixed MCMC *draws*
(not conditional means) from a correct model are marginally distributed as the
prior.  So its pooled SD across subjects and chains targets exactly **1.0** at
every iteration, not just at convergence.  No reference chain and no second run
is needed to interpret it.

**But a pooled SD of 1.0 does not establish that the chain is mixing**, and it
is worth being explicit about that, because the number is reassuring and the
inference from it is wrong.  The SD describes the *marginal spread* of the
draws; a chain that has stopped moving retains whatever spread it already had,
so it reports a perfectly healthy 1.0 forever.  Measured on Bauer's gamma model
at the package defaults, the two declared latents have pooled SDs of 1.055 and
0.970 -- and lag-1 autocorrelations of **0.991 and 0.975**.  The spread is
right and the chain is barely moving.

`mcmcPhiAcf` is the trace that catches that, and on this model it is the one
that matters.  An autocorrelation near 1 means this iteration's draws are
essentially last iteration's, which is exactly the input that collapses a
distribution M-step: refitting a family to the same draws over and over is the
"barely mixing" row of the demonstration above.

Read the two together.  A latent whose pooled SD has drifted below 1 produces a
fitted family that is too tight, which tightens the next iteration's prior --
the fit then reports a confident typical value with no between-subject
variability left.  A latent whose SD is 1 but whose autocorrelation is 0.99 is
the same hazard arriving without the warning sign.

Measured at the package defaults on Bauer's gamma model (300 subjects, means
over the last 20 iterations):

```
acceptance   rw=0.241  coord=0.483        (targets 0.234 and 0.44)
subjects that accepted NOTHING: 36%
```

Both pooled acceptance rates sit essentially on their theoretical optima --
which is what `iaccept` and `iacceptSingle` adapt them to -- while **more than a
third of the population is frozen**.  That is the discrepancy `mcmcStuck` exists
to expose, and it is invisible in every other number the fit reports: a
population rate of 0.24 is equally consistent with every subject at 0.24 and
with a third of them never moving.

This diagnostic earned its place twice over.  It is what revealed that saem's
acceptance adaptation had been freezing the chain **completely** on ordinary
models -- on plain `theo_sd` every subject was frozen and the objective was 60
units worse than `origin/main`, with nothing in the fit reporting it -- and it
is what showed that the first attempt at per-subject adaptation made the problem
worse rather than better.  Both were invisible to the objective, the parameter
estimates and the acceptance rate alike.

## Transcribing Bauer's own control stream

Bauer's `gamma_indpar` SAEM stream is the reference answer for the declared
gamma model, and it is worth reading for what it does *not* do:

```
$OMEGA BLOCK(2) VALUES(0.3,0.01)   ; ETA(1),ETA(2)  Q and V2, real IIV
$OMEGA (1.0 FIXED)X2               ; ETA(3),ETA(4)  the two latent normals
$OMEGA (0.0 FIXED)X5               ; ETA(5)..ETA(9) MU_ helper etas
```

with `MU_5..MU_9` carrying the two gamma means, the two relative variances and
the copula element `L21`.  So:

* the latent normals are **diagonal** unit-variance; the correlation lives in
  `$PK` as `L21*ETA(3) + sqrt(1-L21^2)*ETA(4)`, not in `$OMEGA` -- which is
  what `rxEtaDistExpand()` reproduces;
* **every** distribution parameter, and the copula element, is
  **mu-referenced with a helper random effect of variance exactly zero**.  None
  of them is a non-mu theta.

It is tempting to port that third line directly -- `rxode2::rxEtaDistMuRef()`
creates the helper etas, and `saemControl(zeroOmegaDirect = TRUE)` supplies the
direct route a zero variance needs.  **That has been tried in nlmixr2 and it
makes the fit worse, not better.**  The observation is right and the conclusion
does not carry over, for a reason worth stating:

* NONMEM handles an exactly-zero `$OMEGA` on a mu-referenced eta internally.
* nlmixr2's saem cannot sample a zero-variance random effect at all, so
  `zeroOmegaTune` has to write in a non-zero width -- and a helper eta with a
  real width **is** a real random effect on the gamma mean, on its relative
  variance and on `rho`.  Every subject then gets its own, which absorbs the
  between-subject variability that belongs to the latent normals and leaves
  them under-dispersed.

Neither end of that trade-off works: zero cannot move, non-zero absorbs the
variability.  In nlmixr2 a declared distribution's parameters belong on the
**non-mu** route (`refinePhi0Lik`, the `zeroOmega*` options' `else` branch),
which is what `dist()` gives them by default.  The divergence from NONMEM is
elsewhere -- in the kernels of Difference 1 and the step of Difference 2.

### What that stream actually does, and how long it takes

`gamma_clv1_saem.ctl` has **no** preceding ITS or Laplace step -- SAEM runs
cold from `$THETA ... -3 -3 0.3`.  What it does have is

```
$EST METHOD=SAEM ... NBURN=5000 NITER=500 CTYPE=3
```

**a five-thousand-iteration burn-in.**  The `.ext` trajectory shows where the
work happens (THETA5 and THETA6 are the two gamma relative variances, THETA7 is
the copula element `L21`; all three are mu-referenced onto helper etas with
`$OMEGA (0.0 FIXED)`):

| iteration | THETA1 | THETA5 | THETA6 | THETA7 |
|---|---|---|---|---|
| -5000 (start) | 1.50000 | -3.00000 | -3.00000 | 0.30000 |
| -4000 | 1.59112 | -2.53152 | -2.55892 | 0.42588 |
| 0 (end of burn-in) | 1.56391 | -2.43855 | -2.50978 | 0.36270 |
| 500 (final) | 1.56550 | -2.42126 | -2.52589 | 0.41119 |

Two things follow.  First, **NONMEM does move these thetas** -- THETA5 travels
0.58 -- so "a zero-variance mu-referenced theta is simply pinned" is not a
description of NONMEM's behavior, whatever route it takes to move them.  (The
`.ext` cannot say *which* route; as shown above the mu step's `Omega^-1`
cancels, so it is not degenerate on its face, and the guide scopes eqs.
1.47-1.52 to thetas *not* expressed through mu, which these are.)

Second, essentially all of the movement happens in the first ~1000 burn-in
iterations and the 500 accumulation iterations barely change anything.  A cold
nlmixr2 `saem` run at `nBurn = 100, nEm = 100` is not a shorter version of this
experiment; it is a different one.

## Do they matter?  One measured example

Bauer's gamma model (300 subjects, seed 99, `nBurn = nEm = 100` cold).  The
simulation truth is known, so the last column is the mean absolute relative
error against it rather than a likelihood.

| arm | CL | V1 | rel.var CL | rel.var V1 | cor | MARE | frozen |
|---|---|---|---|---|---|---|---|
| default | 4.448 | 5.158 | 0.0392 | 0.0516 | 0.846 | 42.0% | 36% |
| `rwOmega = TRUE` | 4.447 | 5.170 | 0.0390 | 0.0516 | 0.827 | 41.2% | 36% |
| `etaDistEvery = 20` | 4.668 | 4.532 | 0.0543 | 0.0406 | 0.585 | 26.6% | 13% |
| `iacceptPerId = TRUE` | 4.725 | 5.313 | 0.0824 | 0.0998 | 0.142 | 21.6% | 8% |
| **both of the last two** | 5.269 | 4.835 | **0.0885** | 0.0604 | 0.476 | **10.0%** | 1% |
| ... plus `rwOmega` | 5.264 | 4.835 | 0.0883 | 0.0604 | 0.476 | 9.9% | 1% |
| truth | 5.03 | 4.66 | 0.086 | 0.086 | 0.438 | | |
| NONMEM `METHOD=SAEM` | 4.786 | 4.606 | 0.0888 | -- | 0.411 | ~3.8% | |

The two that matter are complementary rather than redundant, and the mechanism
is visible in the columns:

* `iacceptPerId` fixes the **marginal spreads** -- both relative variances land
  near the truth (0.0824 and 0.0998 against 0.086) -- while driving the
  correlation too far *down* (0.142 against 0.438).
* `etaDistEvery = 20` fixes the **dependence** through the spaced copula M-step
  (0.585) while leaving both variances low.
* Together: 0.0885 and 0.476, and less than half the error of either alone.

`rwOmega` does essentially nothing on this model -- 41.2% against 42.0% alone,
9.9% against 10.0% combined, with every parameter agreeing to three digits.
That is worth stating plainly because it is the difference the comparison above
makes the strongest theoretical case for: NONMEM and Monolix both propose from
`lambda * Omega` and only saemix walks on a diagonal.  On a chain that is
otherwise mixing, the shape of the random-walk proposal turns out not to be
what was limiting this fit.

**These numbers replace an earlier set that was measured against a broken
baseline.**  saem's acceptance-rate adaptation had two defects that froze the
chain (kernels sharing one scale vector adapted toward different targets, and a
permanent NaN latch in the acceptance test); every arm measured before they were
fixed was partly measuring how well a setting bypassed that.  The corrected
figures are smaller and, unlike the originals, decompose sensibly.  The episode
is also the argument for the diagnostics: nothing else in the fit reported that
100% of subjects had stopped moving.

## Summary of levers

| Gap | nlmixr2 setting that closes it | Default |
|---|---|---|
| no mode 1B kernel | `saemControl(nu1B = 2, nb1B = 10)` | off |
| diagonal rather than `Omega`-shaped random walk | `rwOmega = TRUE` | `FALSE` |
| pooled rather than per-subject acceptance adaptation | `iacceptPerId = TRUE` | `FALSE` |
| damped argmax rather than one accepted BHHH step | `nonMuThetaBhhh = TRUE` | `FALSE` |
| GLS cannot move a zero-variance mu theta | `zeroOmegaDirect = TRUE` | `FALSE` |
| exploration width never decays | `zeroOmegaAnneal = 1000` | `0` |
| non-mu refinement starts at half the *total* | `nonMuThetaStart = round(nBurn/2)` | half of `nBurn + nEm` |
| single chain vs several | `nmc` | `3` |

The two `zeroOmega*` rows apply to a model that genuinely has a mu-referenced
random effect with no between-subject variability.  They are **not** a route to
Bauer's arrangement for a declared distribution -- see the section above.

(`nonMuThetaBhhh` carries two of NONMEM's changes at once: the information
matrix becomes the per-subject outer product `sum_i g_i g_i'` rather than the
per-observation one, and the step must pass the `alpha` acceptance test.)

Not closable by an option, and listed so they are not mistaken for tuning:
Monolix's averaging of the *objective* rather than the maximizer, its second
annealing coefficient `tau2` for the residual variance, and the frozen-parameter
burn-in phase (`NBURN` / `K_b` / `nbiter.burn`) that the other three have and
nlmixr2 does not.

## References

* Bauer, R.J. *NONMEM 7 Technical Guide*, sections on the MCMC method of
  Expectation in SAEM (eqs. 1.133-1.153) and the maximization step
  (eqs. 1.45-1.52).
* Bauer, R.J. NONMEM Tutorial Part II: Estimation Methods and Advanced Examples.
  *CPT Pharmacometrics Syst Pharmacol* 2019.
* *Monolix Methodology*, sections 1.1.1-1.1.3 (the SAEM, MCMC-SAEM and
  Simulated Annealing SAEM algorithms).
* Comets, E., Lavielle, M., Mentre, F. Parameter estimation in nonlinear mixed
  effect models using saemix. *J Stat Softw* 2017; and the saemix 3.5 sources.
* Kuhn, E., Lavielle, M. Coupling a stochastic approximation version of EM with
  an MCMC procedure. *ESAIM P&S* 2004.
