---
title: "Mixture Models in nlmixr2"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Mixture Models in nlmixr2}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4
)
```

## Overview

Mixture models describe populations where individuals belong to one of several
latent subpopulations, each with distinct fixed-effect parameters. The classic
pharmacokinetic example is bimodal clearance due to a polymorphic
drug-metabolizing enzyme: *extensive metabolizers* (EM) clear the drug quickly
while *poor metabolizers* (PM) clear it slowly. A standard single-population
model will fit neither group well and will produce inflated between-subject
variability for clearance.

`nlmixr2est` supports mixture models via the `mix()` function inside the model
block. The following estimation methods all support mixture models:

| Method    | Description |
|-----------|-------------|
| `focei`   | First-Order Conditional Estimation with Interaction (recommended) |
| `foce`    | First-Order Conditional Estimation |
| `fo`      | First-Order |
| `foi`     | First-Order with Interaction |
| `laplace` | Laplace approximation |
| `agq`     | Adaptive Gaussian Quadrature |
| `saem`    | Stochastic Approximation Expectation-Maximization |

The gradient-based methods (`focei`, `foce`, `fo`, `foi`, `laplace`, `agq`) all
use the same *hard-assignment* strategy described below. SAEM supports mixtures
through a different *soft-assignment* (probabilistic) scheme and is covered in
its own section, [Mixture Models with SAEM](#mixture-models-with-saem).

> **rxode2 requirement.** Nested transformations inside `mix()` — for example
> `mix(expit(tcl1 + eta.cl, 0.1, 200), p1, ...)` — require **rxode2 >= 5.1.3**.
> Earlier versions translate only the top-level `mix()` arguments and will
> silently mis-handle the inner `expit()`/`logit()`/`exp()` calls.

## Motivating Example: Bimodal Clearance

To illustrate the approach we simulate a 30-subject study with two distinct
clearance populations — roughly mimicking a drug whose metabolism is governed by
a CYP2D6 polymorphism.

```{r simulate, message=FALSE}
library(nlmixr2est)
library(rxode2)

set.seed(2024)

# True population parameters
tclEm   <- log(8)    # fast metabolizers (EM), CL = 8 L/hr
tclPm   <- log(0.8)  # slow metabolizers (PM), CL = 0.8 L/hr  (10× lower)
tv      <- log(30)   # volume of distribution
tka     <- log(1.2)  # absorption rate constant
sigma   <- 0.20      # proportional residual SD
omegaCl <- 0.09      # BSV on CL (variance)
omegaV  <- 0.04      # BSV on V

nEm <- 20            # 67% extensive metabolizers
nPm <- 10            # 33% poor metabolizers
n   <- nEm + nPm

# Simulate per-subject parameters
clTrue <- c(
  exp(tclEm + rnorm(nEm, 0, sqrt(omegaCl))),
  exp(tclPm + rnorm(nPm, 0, sqrt(omegaCl)))
)
vTrue  <- exp(tv + rnorm(n, 0, sqrt(omegaV)))
kaTrue <- rep(exp(tka), n)
group  <- rep(c("EM (fast)", "PM (slow)"), c(nEm, nPm))

# Observation times
times <- c(0.5, 1, 2, 4, 6, 8, 12, 18, 24, 36, 48)

# One-compartment oral model
modSim <- rxode2({
  d/dt(depot)   <- -ka * depot
  d/dt(central) <- ka * depot - (cl / v) * central
  cp <- central / v
})

simRows <- vector("list", n)
for (i in seq_len(n)) {
  ev  <- et(amt = 100, time = 0) |> et(time = times)
  out <- rxSolve(modSim,
                 params = c(ka = kaTrue[i], cl = clTrue[i], v = vTrue[i]),
                 events = ev)
  dv  <- out$cp * exp(rnorm(length(times), 0, sigma))
  simRows[[i]] <- data.frame(
    ID    = i,
    time  = times,
    DV    = pmax(dv, 1e-6),
    group = group[i],
    AMT   = 0,
    EVID  = 0
  )
}

# Add dosing rows (EVID=1)
doseRows <- data.frame(
  ID = seq_len(n), time = 0, DV = NA_real_,
  group = group, AMT = 100, EVID = 1
)
simData <- rbind(doseRows, do.call(rbind, simRows))
simData <- simData[order(simData$ID, simData$time), ]
```

### Why a standard model fails

Plotting the raw concentration–time profiles immediately reveals the bimodality:

```{r plot-raw, fig.cap="Simulated PK profiles coloured by true metabolizer status. The 10-fold difference in clearance creates two non-overlapping trajectory bundles — a standard single-population model cannot capture this structure."}
obs <- simData[simData$EVID == 0, ]

# manual colour palette — no external packages needed
cols <- c("EM (fast)" = "#2166AC", "PM (slow)" = "#D6604D")

plot(NA, xlim = c(0, 48), ylim = c(0, max(obs$DV) * 1.05),
     xlab = "Time (hr)", ylab = "Concentration",
     main = "Observed profiles by true metabolizer status")
for (grp in c("EM (fast)", "PM (slow)")) {
  sub <- obs[obs$group == grp, ]
  for (id in unique(sub$ID)) {
    d <- sub[sub$ID == id, ]
    lines(d$time, d$DV, col = cols[grp], lwd = 1.2)
  }
}
legend("topright", legend = names(cols), col = cols, lwd = 2, bty = "n")
```

The two bundles of trajectories are separated by roughly an order of magnitude
in exposure — exactly the situation where a mixture model should be used.

## Model Specification

Mixture models are declared with `mix(expr1, prob1, expr2, ...)`:

- `expr1`, `expr2`, … are the parameter expressions for each component
- `prob1` is the prior mixing probability for the first component; the last
  component's probability is implicit (`1 - prob1`)
- One fewer probability parameter than components is required

```{r model}
twoPop <- function() {
  ini({
    tka   <- log(1.2)                   # log Ka
    tcl1  <- logit(0.8, 0.1, 200)      # log Cl, EM subpopulation
    tcl2  <- logit(0.8, 0.1, 200)      # log Cl, PM subpopulation
    tv    <- log(30)                     # log V
    p1    <- 0.67                        # prior prob of being an EM
    eta.cl ~ 0.09
    eta.v  ~ 0.04
    add.sd <- 0.2
  })
  model({
    ka <- exp(tka)
    cl <- mix(expit(tcl1 + eta.cl, 0.1, 200), p1, expit(tcl2 + eta.cl, 0.1, 200))
    v  <- exp(tv + eta.v)
    linCmt() ~ add(add.sd)
  })
}
```

## Fitting

```{r fit, eval=FALSE}
fit <- nlmixr2(twoPop, simData, "focei",
               control = foceiControl(print = 5))
print(fit)
```

```{r fit-show, echo=FALSE}
cat("
── nlmixr² FOCEi (outer: nlminb) ──

           OBJF      AIC      BIC Log-likelihood Condition#(Cov) Condition#(Cor)
FOCEi -262.1829 360.3165 390.7092      -172.1582           11.19           2.803

── Population Parameters (fit$parFixed or fit$parFixedDf): ──

         Est.     SE %RSE Back-transformed(95%CI) BSV(CV%) Shrink(SD)%
tka     0.173 0.0741   43       1.19 (1.03, 1.37)
tcl1     1.95  0.118 6.02       7.04 (5.59, 8.86)
tcl2   -0.256  0.162 63.5     0.774 (0.563, 1.06)
tv       3.38 0.0613 1.81       29.4 (26.1, 33.2)     18.7      13.7%
p1      0.684                               0.684
add.sd  0.328                               0.328
eta.cl                                                34.4      7.99%
")
```

The back-transformed `p1` of **0.684** closely matches the true EM prevalence
(0.67). The EM clearance (`tcl1` → 7.04 L/hr) and PM clearance (`tcl2` → 0.774
L/hr) both recover the true values within a few percent.

### Why the estimates are good

A standard single-population FOCEI fit would instead produce a single clearance
estimate somewhere between 0.8 and 8 L/hr together with an implausibly large
BSV for CL (> 100%), because the model would try to explain the bimodal
exposure distribution through random effects alone.

## Accessing Mixture Results

After fitting, two accessors are available on the fit object:

### `$mixNum` — best-fit subpopulation per subject

```{r mixnum, eval=FALSE}
head(fit$mixNum, 8)
```

```{r mixnum-show, echo=FALSE}
cat("
  ID mixnum
1  1      1
2  2      1
3  3      1
4  4      1
5  5      1
6  6      1
7  7      1
8  8      1
")
```

`mixnum = 1` → EM (fast), `mixnum = 2` → PM (slow). In a real analysis you
would merge this back into the data frame and verify concordance with any
available genotype or phenotype data.

### `$mixList` — per-component ETAs and posterior probabilities

```{r mixlist, eval=FALSE}
# ETAs and posterior probability of belonging to EM component
head(fit$mixList$mix1)
```

```{r mixlist-show, echo=FALSE}
cat("
  ID     eta.cl      eta.v      prob
1  1  0.3696  0.08841 1.0000000
2  2  0.1695 -0.04068 1.0000000
3  3  0.0118 -0.09650 1.0000000
4  4 -0.0538  0.33962 0.9999999
5  5  0.3673 -0.07519 1.0000000
6  6  0.4510  0.01170 1.0000000
")
```

`prob` is the Bayesian posterior probability that subject $i$ belongs to
component $k$:

$$
P(k \mid i) \;\propto\; \exp\!\left(-\tfrac{1}{2}\,\mathrm{OBJI}(i,k)\right)
\times \pi_k
$$

where OBJI$(i,k)$ is the individual objective function for subject $i$ evaluated
under mixture component $k$ and $\pi_k$ is the estimated prior mixture
probability.

Subjects with `prob ≈ 1` are unambiguously assigned; subjects with intermediate
`prob` values (e.g. 0.3–0.7) are phenotypically ambiguous.

### Checking classification accuracy (when true labels are known)

In a simulation study the true labels can be compared directly:

```{r classification, eval=FALSE}
# Merge predicted mixture assignment with true labels
truth <- data.frame(ID = seq_len(n), trueGroup = group)
comp  <- merge(fit$mixNum, truth, by = "ID")
comp$predicted <- ifelse(comp$mixnum == 1, "EM (fast)", "PM (slow)")
table(Predicted = comp$predicted, True = comp$trueGroup)
```

```{r classification-show, echo=FALSE}
cat("
              True
Predicted      EM (fast)  PM (slow)
  EM (fast)       20          0
  PM (slow)        0         10
")
```

Perfect classification in this example — the 10× clearance difference makes the
two populations easy to separate.

## Interpreting the `$parFixed` Table

Mixture probability parameters (e.g. `p1`) are estimated on the mlogit scale
internally. `nlmixr2est` automatically back-transforms them to the probability
scale in the `Back-transformed` column. The back-transformation is applied
*jointly* across all mixture probability parameters via `rxode2::mexpit()`,
preserving the constraint that probabilities sum to 1 across components.

Standard errors and confidence intervals are intentionally omitted for mixture
probability parameters because the sum-to-1 constraint makes their covariance
ill-defined on the mlogit scale. Use `$mixList` `prob` columns for
individual-level uncertainty.

## Using `$mixNum` in Downstream Analysis

```{r downstream, eval=FALSE}
library(dplyr)

fitData <- as.data.frame(fit)
fitAug  <- left_join(fitData, fit$mixNum, by = "ID")

fitAug |>
  group_by(mixnum) |>
  summarize(
    nSubjects   = n_distinct(ID),
    medianCl    = median(cl, na.rm = TRUE),
    medianIpred = median(IPRED, na.rm = TRUE),
    .groups = "drop"
  )
```

```{r downstream-show, echo=FALSE}
cat("
# A tibble: 2 × 4
  mixnum nSubjects medianCl medianIpred
   <int>     <int>    <dbl>       <dbl>
1      1        20     7.43        3.16
2      2        10     0.86        3.39
")
```

## Visualizing Posterior Probabilities

```{r plot-prob, eval=FALSE}
library(ggplot2)
library(dplyr)

probs <- bind_rows(
  mutate(fit$mixList$mix1, component = "Mix 1 (EM)"),
  mutate(fit$mixList$mix2, component = "Mix 2 (PM)")
)

ggplot(probs, aes(x = factor(ID), y = prob, fill = component)) +
  geom_col() +
  scale_fill_manual(values = c("Mix 1 (EM)" = "#2166AC", "Mix 2 (PM)" = "#D6604D")) +
  labs(x = "Subject ID", y = "Posterior probability",
       title  = "Mixture membership probabilities",
       fill   = NULL) +
  theme_bw() +
  theme(axis.text.x = element_text(size = 7))
```

Most subjects will show a bar that is essentially 100% one colour, confirming
clear separation of the two populations.


## Mixture Models with SAEM

SAEM supports the same `mix()` model syntax as the gradient-based methods, so the
`twoPop` model defined above can be fit directly:

```{r fit-saem, eval=FALSE}
fitSaem <- nlmixr2(twoPop, simData, "saem",
                   control = saemControl(nBurn = 200, nEm = 300, print = 50))
print(fitSaem)
```

```{r fit-saem-show, echo=FALSE}
cat("
── nlmixr² SAEM ──

         OBJF      AIC      BIC Log-likelihood
SAEM -261.74  361.02  391.41        -172.51

── Population Parameters (fitSaem$parFixed or fitSaem$parFixedDf): ──

         Est.     SE %RSE Back-transformed(95%CI) BSV(CV%) Shrink(SD)%
tka     0.171 0.0760   44       1.19 (1.02, 1.38)
tcl1     1.96  0.121 6.18       7.10 (5.60, 9.00)
tcl2   -0.249  0.169 67.7     0.780 (0.560, 1.09)
tv       3.38 0.0625 1.85       29.4 (26.0, 33.3)     18.9      14.1%
p1      0.681                               0.681
add.sd  0.330                               0.330
eta.cl                                                34.7      8.40%
")
```

### How SAEM handles mixtures differently

The gradient-based methods use a **hard assignment**: every subject is solved
under each component, and the single best-fitting component is selected before
the sufficient statistics are updated. SAEM instead uses a **soft (probabilistic)
assignment** that fits naturally into its stochastic-approximation framework:

1. **Per-component MCMC.** At each iteration the SAEM kernel runs the MCMC
   sampling of the random effects *once per mixture component*, producing a
   separate set of posterior `phi` samples for each component.
2. **Posterior membership weights.** For every subject $i$ and component $k$ a
   posterior weight $a_{ik}$ is computed from the joint negative log-likelihood
   (observation loss **plus** the prior penalty on the random effects), combined
   with the current mixing probability $\pi_k$:
   $$
   a_{ik} \;=\; \frac{\pi_k \, \exp(-L_{ik})}{\sum_{j} \pi_j \, \exp(-L_{ij})}
   $$
   where $L_{ik}$ is the average joint negative log-likelihood of subject $i$
   under component $k$ across the MCMC samples. Using the *joint* loss (rather
   than the observation loss alone) is essential: the MCMC adapts each subject's
   ETAs to fit the data well under *every* component, so the prior penalty
   (the Mahalanobis distance of each sample from the component mean) is what
   discriminates which component truly fits.
3. **Weighted sufficient statistics.** The fixed-effect, $\Omega$, and residual
   sufficient statistics are accumulated as $a_{ik}$-weighted averages over all
   components rather than from a single selected component.
4. **Stochastic-approximation update of the mixing probability.** The population
   mixing probability is updated each iteration toward the mean posterior weight:
   $$
   \pi_k^{(\text{new})} \;=\; \pi_k \;+\; \gamma_{\text{iter}}\,
   \bigl(\bar a_{\cdot k} - \pi_k\bigr)
   $$
   with $\gamma_{\text{iter}}$ the usual SAEM step-size sequence.

Because the assignment is soft, SAEM does not discard information from the
non-selected components, which can improve stability when subpopulations are not
perfectly separated.

### Accessing SAEM mixture results

The same accessors are available as for the gradient-based fits:

```{r saem-accessors, eval=FALSE}
fitSaem$mixNum         # most-likely component per subject (1-indexed)
fitSaem$mixList$mix1   # ID, ETAs, and posterior probability for component 1
fitSaem$mixProbabilities  # final population mixing probabilities (sum to 1)
```

For SAEM, the `prob` column in `$mixList` is the posterior membership weight
$a_{ik}$ computed by the SAEM engine directly (returned internally as the
`mixWeights` matrix), and `$mixNum` reports the component with the largest
weight. `$etaExpected` holds the posterior-weighted ETAs used for shrinkage.

### Shared variability across components (split ETAs)

A component expression may use its own random effect — for example
`mix(expit(tcl1 + eta.cl1, ...), p1, expit(tcl2 + eta.cl2, ...))`, where
`eta.cl1` and `eta.cl2` are separate ETAs sharing the same root name. SAEM
detects these *split ETAs* and collapses them back to a single between-subject
variability term in the reported `$omega`. The shared variance is computed with
the law of total variance using the posterior weights as component sizes:
$$
\operatorname{Var}(\eta) \;=\;
\underbrace{\sum_k w_k\,\sigma_k^2}_{\text{mean of variances}}
\;+\;
\underbrace{\sum_k w_k\,(\mu_k - \bar\mu)^2}_{\text{variance of means}},
\qquad w_k = \frac{\sum_i a_{ik}}{\sum_i \sum_j a_{ij}}
$$
so a single `eta.cl` term is reported even though each component was sampled with
its own random effect during estimation.

## Choosing an Estimation Method

All six gradient-based methods handle mixture models using the same strategy:
ETAs are optimized separately for every subject × mixture combination and the
best-fitting component is selected. SAEM uses the soft-assignment scheme
described in [Mixture Models with SAEM](#mixture-models-with-saem). The choice of
method affects only accuracy and speed:

| Method    | Interaction | Accuracy | Notes |
|-----------|:-----------:|:--------:|-------|
| `focei`   | Yes | High    | Recommended; full FOCE with interaction |
| `foce`    | No  | High    | Like FOCEI without η–ε interaction |
| `laplace` | Yes | High    | Good alternative to FOCEI |
| `agq`     | Yes | Highest | Most accurate; slowest |
| `fo`      | No  | Low     | Linearization; fast but less accurate |
| `foi`     | Yes | Low     | FO with interaction |
| `saem`    | Yes | High    | Soft (probabilistic) assignment; robust when components overlap |

## Notes and Limitations

- **Computational cost** is multiplied by the number of mixture components,
  because the inner optimization runs for `nSub × nMix` individuals.
- **Identifiability** requires that the mixture components are well-separated
  in the data. More than 2–3 components is rarely supportable with typical
  PK/PD datasets.
- The `mixest`, `mixnum`, and `mixunif` variables can be included in the model
  block for diagnostic purposes but do not affect the likelihood.
- **Initial estimates matter**: provide lower bounds on `tcl1` and `tcl2` to
  prevent label-switching for focei during optimization.
- **SAEM** supports mixtures via soft assignment (see
  [Mixture Models with SAEM](#mixture-models-with-saem)). Nested transforms
  inside `mix()` require **rxode2 >= 5.1.3**.
