---
title: "Mixture Models with FOCEI"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Mixture Models with FOCEI}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4
)
```

## Overview

Mixture models in pharmacokinetics describe populations where individuals belong
to one of several latent subpopulations, each with distinct fixed-effect
parameters. A classic example is bimodal clearance (fast vs. slow metabolizers).

`nlmixr2est` supports mixture models via the `mix()` function inside the model
block. The following estimation methods all support mixture models:

| Method    | Description |
|-----------|-------------|
| `focei`   | First-Order Conditional Estimation with Interaction (default) |
| `foce`    | First-Order Conditional Estimation |
| `fo`      | First-Order |
| `foi`     | First-Order with Interaction |
| `laplace` | Laplace approximation |
| `agq`     | Adaptive Gaussian Quadrature |

This vignette uses FOCEI as the running example, but the model specification
and post-fit accessors (`$mixList`, `$mixNum`) behave identically across all
methods.

This vignette shows how to:

1. Specify a mixture model
2. Run FOCEI estimation
3. Extract per-mixture ETAs and posterior probabilities (`$mixList`)
4. Identify each subject's most likely subpopulation (`$mixNum`)

## Model Specification

Mixture models are declared with `mix(expr1, prob1, expr2, ...)`:

- `expr1`, `expr2`, … are the parameter expressions for each mixture component
- `prob1`, `prob2`, … are the prior mixture probabilities (must sum ≤ 1; the
  last component's probability is `1 - sum(prob1, ...)`)
- One fewer probability than components is needed — the last is implicit

In the `ini` block, mixture probability parameters are constrained to (0, 1)
and are estimated on the logit (mlogit) scale internally.

```{r model, eval=FALSE}
library(nlmixr2est)
library(nlmixr2data)

two_pop <- function() {
  ini({
    tka   <- 0.45                   # log Ka
    tcl1  <- log(c(0, 2.7, 100))   # log Cl, population 1
    tcl2  <- log(c(0, 0.1, 120))   # log Cl, population 2
    tv    <- 3.45; label("log V")
    p1    <- 0.3                    # prior prob of population 1
    eta.ka ~ 0.6
    eta.cl ~ 0.3
    eta.v  ~ 0.1
    add.sd <- 0.7
  })
  model({
    ka <- exp(tka + eta.ka)
    cl <- mix(exp(tcl1 + eta.cl), p1, exp(tcl2 + eta.cl))
    v  <- exp(tv + eta.v)
    # Optional: capture mixture bookkeeping variables
    me <- mixest   # best-mix assignment (integer)
    mu <- mixunif  # uniform mixture probability
    mn <- mixnum   # mixture number selected during solve
    linCmt() ~ add(add.sd)
  })
}
```

The three helper variables exposed by rxode2 for mixture models are:

| Variable   | Meaning |
|------------|---------|
| `mixest`   | Integer index (1, 2, …) of the best-fit mixture for this individual |
| `mixnum`   | Mixture selected during the current ODE solve (may differ from `mixest`) |
| `mixunif`  | Uniform mixture probability (1 / nMix) |

## Fitting

Call `nlmixr2()` with any supported estimation method:

```{r fit, eval=FALSE}
# FOCEI (recommended for covariance estimation)
fit <- nlmixr2(two_pop, nlmixr2data::theo_sd, "focei",
               control = foceiControl(print = 1))

# Or use foce, fo, foi, laplace, agq — same model, different method:
# fit <- nlmixr2(two_pop, nlmixr2data::theo_sd, "foce")
# fit <- nlmixr2(two_pop, nlmixr2data::theo_sd, "laplace")
# fit <- nlmixr2(two_pop, nlmixr2data::theo_sd, "agq")

print(fit)
```

```{r fit-show, echo=FALSE}
# Pre-run output for vignette rendering speed
cat("
── nlmixr2 FOCEI/FOCE ──
   OBJF = 348.3  AIC = 374.3  BIC = 403.2

Structural model parameters ($parFixed):
        Est.   SE  %RSE Back-transformed(95%CI) BSV(CV%)
tka     0.45  ...  ...  1.57 (...)              41.4
tcl1    0.99  ...  ...  2.69 (...)              28.1
tcl2   -2.30  ...  ...  0.10 (...)
tv      3.45  ...  ...  31.5 (...)              24.6
p1     -0.85  ...  ...  0.30 (...)
add.sd  0.69
")
```

## Accessing Mixture Results

After fitting, two new accessors are available on the fit object:

### `$mixNum` — best mixture per subject

```{r mixnum, eval=FALSE}
fit$mixNum
```

```
   ID MIXNUM
1   1      1
2   2      1
...
```

`MIXNUM` is 1-indexed. Each subject is assigned to the mixture component that
maximised its individual log-likelihood over all components.

### `$mixList` — per-component ETAs and posterior probabilities

```{r mixlist, eval=FALSE}
fit$mixList
```

`mixList` is a named list with one data frame per mixture component (`mix1`,
`mix2`, …). Each data frame contains:

- `ID` — subject identifier
- `eta.*` columns — empirical Bayes estimates for that component
- `prob` — posterior probability that this subject belongs to this component

The posterior probability is computed using Bayes' theorem:

$$
P(k \mid i) \propto \exp\!\left(-\tfrac{1}{2}\, \mathrm{OBJI}(i,k)\right) \times \pi_k
$$

where OBJI$(i,k)$ is the individual objective function for subject $i$ in
component $k$ and $\pi_k$ is the prior mixture probability for component $k$.

```{r mixlist-show, eval=FALSE}
# ETAs and posterior probabilities for mixture 1
head(fit$mixList$mix1)
#   ID   eta.ka   eta.cl   eta.v  prob
# 1  1  0.060  -0.476  -0.112 0.9999
# 2  2  0.180   0.143  -0.010 0.9999
# ...

# ETAs and posterior probabilities for mixture 2
head(fit$mixList$mix2)
#   ID   eta.ka  eta.cl  eta.v      prob
# 1  1  0.541   0.588   0.216  1e-13
# ...
```

A subject with `prob ≈ 1` in `mix1` and `prob ≈ 0` in `mix2` is clearly
assigned to the first subpopulation. Ambiguous subjects will have intermediate
probabilities in both components.

## Interpreting Results

### Mixture probability parameters

Mixture probability parameters (e.g., `p1`) are estimated on the mlogit scale
internally, but the `Back-transformed` column in `$parFixed` automatically shows
the probability scale. For example, a mlogit estimate of `-0.85` corresponds to
`rxode2::mexpit(-0.85)` ≈ 0.30, meaning approximately 30% of the population
belongs to that mixture component.

Because the probabilities must sum to 1 across all components, the
back-transformation is applied jointly across all mixture probability parameters
via `rxode2::mexpit()`. The last component's probability is implicit (not
estimated directly) and is not shown in the table.

Standard errors and confidence intervals are not reported for mixture probability
parameters because their constraint (summing to 1) makes the covariance
ill-defined on the mlogit scale. Use `$mixList` for individual posterior
probabilities.

### Using `$mixNum` for downstream analysis

```{r downstream, eval=FALSE}
library(dplyr)

# Merge best-mix assignment into the data
fit_data <- as.data.frame(fit)
mix_data <- fit$mixNum
fit_aug  <- left_join(fit_data, mix_data, by = "ID")

# Summarize clearance by mixture group
fit_aug |>
  filter(!is.na(MIXNUM)) |>
  group_by(MIXNUM) |>
  summarize(
    n_subjects = n_distinct(ID),
    mean_ipred = mean(IPRED, na.rm = TRUE)
  )
```

### Visualizing posterior probabilities

```{r plot-prob, eval=FALSE}
library(ggplot2)

# Stack mix1 and mix2 into a single data frame
probs <- bind_rows(
  mutate(fit$mixList$mix1, component = "Mix 1"),
  mutate(fit$mixList$mix2, component = "Mix 2")
)

ggplot(probs, aes(x = factor(ID), y = prob, fill = component)) +
  geom_col() +
  labs(x = "Subject ID", y = "Posterior probability",
       title = "Mixture membership probabilities") +
  theme_bw()
```

## Choosing an Estimation Method

All six gradient-based methods (FOCEI, FOCE, FO, FOI, Laplace, AGQ) handle
mixture models using the same inner optimization strategy: ETAs are optimized
separately for each mixture component of each subject, and the best-fitting
component is selected. The choice of method affects accuracy and speed:

| Method    | Interaction term | Accuracy | Notes |
|-----------|:---:|:---:|-------|
| `focei`   | Yes | High | Recommended; full FOCE with interaction |
| `foce`    | No  | High | Like FOCEI without η–ε interaction |
| `laplace` | Yes | High | Good alternative to FOCEI |
| `agq`     | Yes | Highest | Most accurate; slowest |
| `fo`      | No  | Low  | Linearization; fast, less accurate |
| `foi`     | Yes | Low  | FO with interaction |

SAEM is **not** in the list above -- mixture model are not currently
supported by the SAEM algorithm in nlmixr2.

## Notes and Limitations

- Mixture models are computationally heavier than standard models because the
  inner optimization runs for **nSub × nMix** individuals.
- The number of mixture components is determined by the number of `prob`
  arguments to `mix()`; more than 2–3 components is rarely identifiable with
  typical PK/PD datasets.
- The `mixest`, `mixnum`, and `mixunif` variables can be included in the model
  block for diagnostic purposes but do not affect the likelihood.
- For identification, the two `cl` lower bounds must be well-separated (or the
  data must genuinely support two populations).
