---
title: "Mixture Models in nlmixr2"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Mixture Models in nlmixr2}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4
)
```

## Overview

Mixture models describe populations where individuals belong to one of several
latent subpopulations, each with distinct fixed-effect parameters. The classic
pharmacokinetic example is bimodal clearance due to a polymorphic
drug-metabolizing enzyme: *extensive metabolizers* (EM) clear the drug quickly
while *poor metabolizers* (PM) clear it slowly. A standard single-population
model will fit neither group well and will produce inflated between-subject
variability for clearance.

`nlmixr2est` supports mixture models via the `mix()` function inside the model
block. The following estimation methods all support mixture models:

| Method    | Description |
|-----------|-------------|
| `focei`   | First-Order Conditional Estimation with Interaction (recommended) |
| `foce`    | First-Order Conditional Estimation |
| `fo`      | First-Order |
| `foi`     | First-Order with Interaction |
| `laplace` | Laplace approximation |
| `agq`     | Adaptive Gaussian Quadrature |
| `saem`    | Stochastic Approximation Expectation-Maximization |

The gradient-based methods (`focei`, `foce`, `fo`, `foi`, `laplace`, `agq`) all
use the same *hard-assignment* strategy described below and reliably separate
mixture components with high precision. SAEM supports mixtures through a
different *soft-assignment* (probabilistic) scheme, described in its own
section, [Mixture Models with SAEM](#mixture-models-with-saem); it is stabilized
against a mixing-probability collapse failure mode by default, but with default
settings may separate components less precisely than the gradient-based
methods on small or weakly-separated datasets. **`focei` is recommended when
precise per-subject classification matters; SAEM is a usable alternative that
may need tuning (see its section) to match it.**

> **rxode2 requirement.** Nested transformations inside `mix()` — for example
> `mix(expit(tcl1 + eta.cl, 0.1, 200), p1, ...)` — require **rxode2 >= 5.1.3**.
> Earlier versions translate only the top-level `mix()` arguments and will
> silently mis-handle the inner `expit()`/`logit()`/`exp()` calls.

## Motivating Example: Bimodal Clearance

To illustrate the approach we simulate a 30-subject study with two distinct
clearance populations — roughly mimicking a drug whose metabolism is governed by
a CYP2D6 polymorphism.

```{r simulate, message=FALSE}
library(nlmixr2est)
library(rxode2)

set.seed(2024)

# True population parameters
tclEm   <- log(8)    # fast metabolizers (EM), CL = 8 L/hr
tclPm   <- log(0.8)  # slow metabolizers (PM), CL = 0.8 L/hr  (10× lower)
tv      <- log(30)   # volume of distribution
tka     <- log(1.2)  # absorption rate constant
sigma   <- 0.20      # proportional residual SD
omegaCl <- 0.09      # BSV on CL (variance)
omegaV  <- 0.04      # BSV on V

nEm <- 20            # 67% extensive metabolizers
nPm <- 10            # 33% poor metabolizers
n   <- nEm + nPm

# Simulate per-subject parameters
clTrue <- c(
  exp(tclEm + rnorm(nEm, 0, sqrt(omegaCl))),
  exp(tclPm + rnorm(nPm, 0, sqrt(omegaCl)))
)
vTrue  <- exp(tv + rnorm(n, 0, sqrt(omegaV)))
kaTrue <- rep(exp(tka), n)
group  <- rep(c("EM (fast)", "PM (slow)"), c(nEm, nPm))

# Observation times
times <- c(0.5, 1, 2, 4, 6, 8, 12, 18, 24, 36, 48)

# One-compartment oral model
modSim <- rxode2({
  d/dt(depot)   <- -ka * depot
  d/dt(central) <- ka * depot - (cl / v) * central
  cp <- central / v
})

simRows <- vector("list", n)
for (i in seq_len(n)) {
  ev  <- et(amt = 100, time = 0) |> et(time = times)
  out <- rxSolve(modSim,
                 params = c(ka = kaTrue[i], cl = clTrue[i], v = vTrue[i]),
                 events = ev)
  dv  <- out$cp * exp(rnorm(length(times), 0, sigma))
  simRows[[i]] <- data.frame(
    ID    = i,
    time  = times,
    DV    = pmax(dv, 1e-6),
    group = group[i],
    AMT   = 0,
    EVID  = 0
  )
}

# Add dosing rows (EVID=1)
doseRows <- data.frame(
  ID = seq_len(n), time = 0, DV = NA_real_,
  group = group, AMT = 100, EVID = 1
)
simData <- rbind(doseRows, do.call(rbind, simRows))
simData <- simData[order(simData$ID, simData$time), ]
```

### Why a standard model fails

Plotting the raw concentration–time profiles immediately reveals the bimodality:

```{r plot-raw, fig.cap="Simulated PK profiles coloured by true metabolizer status. The 10-fold difference in clearance creates two non-overlapping trajectory bundles — a standard single-population model cannot capture this structure."}
obs <- simData[simData$EVID == 0, ]

# manual colour palette — no external packages needed
cols <- c("EM (fast)" = "#2166AC", "PM (slow)" = "#D6604D")

plot(NA, xlim = c(0, 48), ylim = c(0, max(obs$DV) * 1.05),
     xlab = "Time (hr)", ylab = "Concentration",
     main = "Observed profiles by true metabolizer status")
for (grp in c("EM (fast)", "PM (slow)")) {
  sub <- obs[obs$group == grp, ]
  for (id in unique(sub$ID)) {
    d <- sub[sub$ID == id, ]
    lines(d$time, d$DV, col = cols[grp], lwd = 1.2)
  }
}
legend("topright", legend = names(cols), col = cols, lwd = 2, bty = "n")
```

The two bundles of trajectories are separated by roughly an order of magnitude
in exposure — exactly the situation where a mixture model should be used.

## Model Specification

Mixture models are declared with `mix(expr1, prob1, expr2, ...)`:

- `expr1`, `expr2`, … are the parameter expressions for each component
- `prob1` is the prior mixing probability for the first component; the last
  component's probability is implicit (`1 - prob1`)
- One fewer probability parameter than components is required

```{r model}
twoPop <- function() {
  ini({
    tka   <- log(1.2)                   # log Ka
    tcl1  <- logit(0.8, 0.1, 200)      # log Cl, EM subpopulation
    tcl2  <- logit(0.8, 0.1, 200)      # log Cl, PM subpopulation
    tv    <- log(30)                     # log V
    p1    <- 0.67                        # prior prob of being an EM
    eta.cl ~ 0.09
    eta.v  ~ 0.04
    add.sd <- 0.2
  })
  model({
    ka <- exp(tka)
    cl <- mix(expit(tcl1 + eta.cl, 0.1, 200), p1, expit(tcl2 + eta.cl, 0.1, 200))
    v  <- exp(tv + eta.v)
    linCmt() ~ add(add.sd)
  })
}
```

## Fitting

```{r fit, eval=FALSE}
fit <- nlmixr2(twoPop, simData, "focei",
               control = foceiControl(print = 5))
print(fit)
```

```{r fit-show, echo=FALSE}
cat("
── nlmixr² FOCEi (outer: bobyqa) ──

           OBJF      AIC      BIC Log-likelihood Condition#(Cov)
FOCEi -262.2443 360.2551 390.6479      -172.1276        14.14459
      Condition#(Cor)
FOCEi        2.891855

── Population Parameters (fit$parFixed or fit$parFixedDf): ──

         Est.      SE  %RSE Back-transformed(95%CI) BSV(CV% or SD) Shrink(SD)%
tka    0.1712 0.07434 43.42    1.187 (1.026, 1.373)
tcl1   -3.323  0.1272 3.827    7.057 (5.564, 8.939)
tcl2   -5.693  0.1783 3.133  0.7715 (0.5739, 1.051)
tv       3.38 0.06005 1.776    29.38 (26.11, 33.05)          18.00     11.97%
p1     0.6666                                0.6666
add.sd 0.3279                                0.3279
eta.cl                                                        0.36      8.322%
")
```

The back-transformed `p1` of **0.667** closely matches the true EM prevalence
(0.67). The EM clearance (`tcl1` → 7.06 L/hr) and PM clearance (`tcl2` → 0.772
L/hr) both recover the true values within a few percent.

### Why the estimates are good

A standard single-population FOCEI fit would instead produce a single clearance
estimate somewhere between 0.8 and 8 L/hr together with an implausibly large
BSV for CL (> 100%), because the model would try to explain the bimodal
exposure distribution through random effects alone.

## Accessing Mixture Results

After fitting, two accessors are available on the fit object:

### `$mixNum` — best-fit subpopulation per subject

```{r mixnum, eval=FALSE}
head(fit$mixNum, 8)
```

```{r mixnum-show, echo=FALSE}
cat("
  ID mixnum
1  1      1
2  2      1
3  3      1
4  4      1
5  5      1
6  6      1
7  7      1
8  8      1
")
```

`mixnum = 1` → EM (fast), `mixnum = 2` → PM (slow). In a real analysis you
would merge this back into the data frame and verify concordance with any
available genotype or phenotype data.

### `$mixList` — per-component ETAs and posterior probabilities

```{r mixlist, eval=FALSE}
# ETAs and posterior probability of belonging to EM component
head(fit$mixList$mix1)
```

```{r mixlist-show, echo=FALSE}
cat("
  ID      eta.cl       eta.v      prob
1  1  0.39326459  0.08586619 1.0000000
2  2  0.17811121 -0.03913940 1.0000000
3  3  0.01023984 -0.09369114 1.0000000
4  4 -0.05653866  0.33189804 0.9999999
5  5  0.38879404 -0.07281481 1.0000000
6  6  0.47960137  0.01132721 1.0000000
")
```

`prob` is the Bayesian posterior probability that subject $i$ belongs to
component $k$:

$$
P(k \mid i) \;\propto\; \exp\!\left(-\tfrac{1}{2}\,\mathrm{OBJI}(i,k)\right)
\times \pi_k
$$

where OBJI$(i,k)$ is the individual objective function for subject $i$ evaluated
under mixture component $k$ and $\pi_k$ is the estimated prior mixture
probability.

Subjects with `prob ≈ 1` are unambiguously assigned; subjects with intermediate
`prob` values (e.g. 0.3–0.7) are phenotypically ambiguous.

### Checking classification accuracy (when true labels are known)

In a simulation study the true labels can be compared directly:

```{r classification, eval=FALSE}
# Merge predicted mixture assignment with true labels
truth <- data.frame(ID = seq_len(n), trueGroup = group)
comp  <- merge(fit$mixNum, truth, by = "ID")
comp$predicted <- ifelse(comp$mixnum == 1, "EM (fast)", "PM (slow)")
table(Predicted = comp$predicted, True = comp$trueGroup)
```

```{r classification-show, echo=FALSE}
cat("
              True
Predicted      EM (fast)  PM (slow)
  EM (fast)       20          0
  PM (slow)        0         10
")
```

Perfect classification in this example — the 10× clearance difference makes the
two populations easy to separate.

## Interpreting the `$parFixed` Table

Mixture probability parameters (e.g. `p1`) are estimated on the mlogit scale
internally. `nlmixr2est` automatically back-transforms them to the probability
scale in the `Back-transformed` column. The back-transformation is applied
*jointly* across all mixture probability parameters via `rxode2::mexpit()`,
preserving the constraint that probabilities sum to 1 across components.

Standard errors and confidence intervals are intentionally omitted for mixture
probability parameters because the sum-to-1 constraint makes their covariance
ill-defined on the mlogit scale. Use `$mixList` `prob` columns for
individual-level uncertainty.

## Using `$mixNum` in Downstream Analysis

```{r downstream, eval=FALSE}
library(dplyr)

fitData <- as.data.frame(fit)
fitAug  <- left_join(fitData, fit$mixNum, by = "ID")

fitAug |>
  group_by(mixnum) |>
  summarize(
    nSubjects   = n_distinct(ID),
    medianCl    = median(cl, na.rm = TRUE),
    medianIpred = median(IPRED, na.rm = TRUE),
    .groups = "drop"
  )
```

```{r downstream-show, echo=FALSE}
cat("
# A tibble: 2 × 4
  mixnum nSubjects medianCl medianIpred
   <int>     <int>    <dbl>       <dbl>
1      1        20     7.43        3.16
2      2        10     0.86        3.39
")
```

## Visualizing Posterior Probabilities

```{r plot-prob, eval=FALSE}
library(ggplot2)
library(dplyr)

probs <- bind_rows(
  mutate(fit$mixList$mix1, component = "Mix 1 (EM)"),
  mutate(fit$mixList$mix2, component = "Mix 2 (PM)")
)

ggplot(probs, aes(x = factor(ID), y = prob, fill = component)) +
  geom_col() +
  scale_fill_manual(values = c("Mix 1 (EM)" = "#2166AC", "Mix 2 (PM)" = "#D6604D")) +
  labs(x = "Subject ID", y = "Posterior probability",
       title  = "Mixture membership probabilities",
       fill   = NULL) +
  theme_bw() +
  theme(axis.text.x = element_text(size = 7))
```

Most subjects will show a bar that is essentially 100% one colour, confirming
clear separation of the two populations.


## Mixture Models with SAEM

SAEM accepts the same `mix()` model syntax as the gradient-based methods, so the
`twoPop` model defined above can be submitted to it directly:

```{r fit-saem, eval=FALSE}
fitSaem <- nlmixr2(twoPop, simData, "saem",
                   control = saemControl(nBurn = 200, nEm = 300, print = 50))
print(fitSaem)
```

```{r fit-saem-show, echo=FALSE}
cat("
── nlmixr² SAEM OBJF by FOCEi approximation ──

 Gaussian/Laplacian Likelihoods: AIC() or $objf etc.
 FOCEi CWRES & Likelihoods: addCwres()

── Population Parameters (fitSaem$parFixed or fitSaem$parFixedDf): ──

        Est.   SE %RSE Back-transformed(95%CI) BSV(CV% or SD) Shrink(SD)%
tka    0.188 -nan  nan       1.21 (-nan, -nan)
tcl1   -4.09 -nan  nan         3.38 (nan, nan)
tcl2   -6.35 -nan  nan         0.45 (nan, nan)
tv      3.39 -nan  nan       29.6 (-nan, -nan)          18.01     10.97%
p1     0.691                             0.691
add.sd 0.326                             0.326
eta.cl                                                   1.11     23.20%

  Information about run found ($runInfo):
   • error calculating tables, returning without table step
   • diag(V) had non-positive or NA entries; the non-finite result may be dubious
   • NaNs produced
")
```

The back-transformed `p1` of **0.691** is in the right neighborhood of the true
EM prevalence (0.67), and `fitSaem$mixNum` shows a real (not collapsed) split:

```{r mixnum-saem-table, echo=FALSE}
cat("
 1  2
26  4
")
```

Cross-tabulated against the true labels, SAEM gets every true EM subject right
but only 4 of the 10 true PM subjects — noticeably less precise than FOCEI's
perfect separation on the same data:

```{r classification-saem, echo=FALSE}
cat("
           True
Predicted   EM (fast) PM (slow)
  EM (fast)        20         6
  PM (slow)         0         4
")
```

> **Two known limitations, both distinct from component separation itself.**
> First, standard errors above are `NaN`: the linearized Fisher-information
> calculation for this *single, shared* `eta.cl` (used by both components)
> remains ill-conditioned when few subjects are confidently assigned to the
> minority component — the package's own test suite works around exactly this
> by setting `covMethod = 0L` for this configuration. Give each component its
> own random effect instead (see [Shared variability across
> components](#shared-variability-across-components-split-etas) below) and
> `covMethod="linFim"` produces finite SEs normally. Second, separation
> quality itself is sensitive to `nBurn`/`nEm` and to the
> `mixProbPriorN`/`mixProbMethod` settings described next — increasing
> `mixProbPriorN` to 30 and using `nBurn = 400, nEm = 800` recovers a 21/9
> split (`p1` ≈ 0.675) on this same simulated data, much closer to FOCEI. For
> a first look at a mixture model, or when precise per-subject classification
> matters, `focei` remains the more reliable default; SAEM is a usable
> alternative that may need more iterations and/or a higher `mixProbPriorN`
> to match it.

### Why SAEM needs help separating mixture components

The gradient-based methods use a **hard assignment**: every subject is solved
under each component, and the single best-fitting component is selected before
the sufficient statistics are updated — this optimizes each subject
independently on every iteration, so it cannot get stuck. SAEM instead uses a
**soft (probabilistic) assignment** that fits naturally into its
stochastic-approximation framework, but that same stochastic-approximation
mechanism makes the *mixing probability* itself prone to a runaway
positive-feedback collapse: the per-subject responsibility used to update it
is weighted by the *current* mixing probability, so a small, noisy drift
shrinks the responsibility of the shrinking component for every subject (not
just the ones that truly belong to the other component), which shrinks the
next iteration's estimate further. Left unchecked this can drive the mixing
probability all the way to exactly 0 or 1 well before the fixed effects have
separated — `saemControl()`'s `mixProbMethod` (below) exists specifically to
prevent that.

### Stabilizing the mixing-probability estimate: `mixProbMethod`

`saemControl()` exposes two independent ways to stabilize the mixing
probability against the collapse described above:

* **`mixProbMethod = "regularized"` (default).** Blend `mixProbPriorN`
  pseudo-subjects, distributed according to the model's *initial* mixing
  probability (from `ini()`), into the responsibility average used at every
  iteration — equivalent to a Dirichlet-prior/MAP-EM correction on the mixing
  proportions. This changes what the estimate converges toward (pulled toward
  the initial guess, with the pull's relative strength shrinking as the
  number of subjects grows), in exchange for reliably preventing collapse.
  Larger `mixProbPriorN` and more `nBurn`/`nEm` iterations both improve
  separation quality, as shown above.

* **`mixProbMethod = "annealed"`.** Give the mixing-probability update its own
  step-size schedule that decays from iteration 1 (`mixProbStepExp`), instead
  of the shared burn-in step of 1 used for ordinary parameters. This only
  slows *how fast* the estimate can move, without biasing what it converges
  to — but for a systematic (rather than purely noise-driven) collapse,
  slowing the walk to the same wrong answer does not, by itself, prevent
  reaching it. Prefer this when you want to avoid any bias toward the initial
  mixing probability and believe the components are already reasonably
  separated.

```{r mixprob-method, eval=FALSE}
# more separation power, at the cost of more iterations and some bias
# toward the ini()-specified p1:
saemControl(mixProbMethod = "regularized", mixProbPriorN = 30,
            nBurn = 400, nEm = 800)
```

### How SAEM handles mixtures differently

The gradient-based methods use a **hard assignment**: every subject is solved
under each component, and the single best-fitting component is selected before
the sufficient statistics are updated. SAEM instead uses a **soft (probabilistic)
assignment** that fits naturally into its stochastic-approximation framework:

1. **Per-component MCMC.** At each iteration the SAEM kernel runs the MCMC
   sampling of the random effects *once per mixture component*, producing a
   separate set of posterior `phi` samples for each component.
2. **Posterior membership weights.** For every subject $i$ and component $k$ a
   posterior weight $a_{ik}$ is computed from the joint negative log-likelihood
   (observation loss **plus** the prior penalty on the random effects), combined
   with the current mixing probability $\pi_k$:
   $$
   a_{ik} \;=\; \frac{\pi_k \, \exp(-L_{ik})}{\sum_{j} \pi_j \, \exp(-L_{ij})}
   $$
   where $L_{ik}$ is the average joint negative log-likelihood of subject $i$
   under component $k$ across the MCMC samples. Using the *joint* loss (rather
   than the observation loss alone) is essential: the MCMC adapts each subject's
   ETAs to fit the data well under *every* component, so the prior penalty
   (the Mahalanobis distance of each sample from the component mean) is what
   discriminates which component truly fits.
3. **Weighted sufficient statistics.** The fixed-effect, $\Omega$, and residual
   sufficient statistics are accumulated as $a_{ik}$-weighted averages over all
   components rather than from a single selected component.
4. **Stochastic-approximation update of the mixing probability.** The population
   mixing probability is updated each iteration toward the mean posterior weight:
   $$
   \pi_k^{(\text{new})} \;=\; \pi_k \;+\; \gamma_{\text{iter}}\,
   \bigl(\bar a_{\cdot k} - \pi_k\bigr)
   $$
   with $\gamma_{\text{iter}}$ the usual SAEM step-size sequence.

Because the assignment is soft, SAEM does not discard information from the
non-selected components, which can improve stability when subpopulations are not
perfectly separated.

### Accessing SAEM mixture results

The same accessors are available as for the gradient-based fits, and return
values without error:

```{r saem-accessors, eval=FALSE}
fitSaem$mixNum         # most-likely component per subject (1-indexed)
fitSaem$mixList$mix1   # ID, ETAs, and posterior probability for component 1
fitSaem$mixProbabilities  # final population mixing probabilities (sum to 1)
```

For SAEM, the `prob` column in `$mixList` is the posterior membership weight
$a_{ik}$ computed by the SAEM engine directly (returned internally as the
`mixWeights` matrix), and `$mixNum` reports the component with the largest
weight. `$etaExpected` holds the posterior-weighted ETAs used for shrinkage.
As shown above, these are meaningful once `mixProbMethod` is stabilizing the
fit, but their precision (how many subjects land in the correct component)
depends on `mixProbPriorN` and the number of iterations, more so than for the
gradient-based methods.

### Shared variability across components (split ETAs)

A component expression may use its own random effect — for example
`mix(expit(tcl1 + eta.cl1, ...), p1, expit(tcl2 + eta.cl2, ...))`, where
`eta.cl1` and `eta.cl2` are separate ETAs sharing the same root name. SAEM
detects these *split ETAs* and collapses them back to a single between-subject
variability term in the reported `$omega`. The shared variance is computed with
the law of total variance using the posterior weights as component sizes:
$$
\operatorname{Var}(\eta) \;=\;
\underbrace{\sum_k w_k\,\sigma_k^2}_{\text{mean of variances}}
\;+\;
\underbrace{\sum_k w_k\,(\mu_k - \bar\mu)^2}_{\text{variance of means}},
\qquad w_k = \frac{\sum_i a_{ik}}{\sum_i \sum_j a_{ij}}
$$
so a single `eta.cl` term is reported even though each component was sampled with
its own random effect during estimation. This calculation is only as good as
the posterior weights $a_{ik}$ feeding it, so it benefits from the same
`mixProbMethod` stabilization described above, and split ETAs are what makes
`covMethod = "linFim"` produce finite SEs.

> **A separate, still-open issue with split ETAs.** In testing, split-ETA
> mixture fits can converge with `p1` looking reasonable but with `tcl1` and
> `tcl2` themselves settling on nearly *identical* values — i.e. the
> components' central tendencies merge even though the mixing-probability
> collapse described above is fixed. If your split-ETA fit's back-transformed
> `tcl1`/`tcl2` (or other per-component fixed effects) come out suspiciously
> close together, check the fit carefully rather than trusting `$mixNum`/`$omega`
> at face value — this looks like a distinct problem from the mixing-probability
> collapse and is not addressed by `mixProbMethod`.

## Choosing an Estimation Method

All six gradient-based methods handle mixture models using the same strategy:
ETAs are optimized separately for every subject × mixture combination and the
best-fitting component is selected. This works reliably in practice and gives
the cleanest separation. SAEM uses the soft-assignment scheme described in
[Mixture Models with SAEM](#mixture-models-with-saem); with the default
`mixProbMethod = "regularized"` it no longer collapses, but by default it may
separate components less precisely than the gradient-based methods on
small/weakly-separated datasets unless `mixProbPriorN` and `nBurn`/`nEm` are
increased (see that section for a worked example):

| Method    | Interaction | Accuracy | Notes |
|-----------|:-----------:|:--------:|-------|
| `focei`   | Yes | High    | Recommended; full FOCE with interaction |
| `foce`    | No  | High    | Like FOCEI without η–ε interaction |
| `laplace` | Yes | High    | Good alternative to FOCEI |
| `agq`     | Yes | Highest | Most accurate; slowest |
| `fo`      | No  | Low     | Linearization; fast but less accurate |
| `foi`     | Yes | Low     | FO with interaction |
| `saem`    | Yes | Moderate–High | Soft assignment; tune `mixProbPriorN`/`nBurn`/`nEm` for closer-to-FOCEI separation; see [Mixture Models with SAEM](#mixture-models-with-saem) |

## Notes and Limitations

- **Computational cost** is multiplied by the number of mixture components,
  because the inner optimization runs for `nSub × nMix` individuals (or, for
  SAEM, `nMix` MCMC passes per iteration).
- **Identifiability** requires that the mixture components are well-separated
  in the data. More than 2–3 components is rarely supportable with typical
  PK/PD datasets.
- The `mixest`, `mixnum`, and `mixunif` variables can be included in the model
  block for diagnostic purposes but do not affect the likelihood.
- **Initial estimates matter**: provide lower bounds on `tcl1` and `tcl2` to
  prevent label-switching for focei during optimization.
- **SAEM** supports mixtures via soft assignment, stabilized against collapse
  by `saemControl(mixProbMethod=)` (see
  [Mixture Models with SAEM](#mixture-models-with-saem)); it typically needs
  more tuning (iterations, `mixProbPriorN`) than the gradient-based methods to
  match their separation precision, and does not currently support covariance
  for mixtures with a single shared ETA (use split ETAs for that). Nested
  transforms inside `mix()` require **rxode2 >= 5.1.3** regardless of
  estimation method.
